---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
devtools::load_all(".", quiet = TRUE)
```

# LEVAiM

<!-- badges: start -->
<!-- badges: end -->

LEVAiM is an R package for organizing longitudinal multi-assay microbiome data
and fitting trajectory models to features that change over time. The current
working path focuses on a spline/GAM backend: dataset -> model specification ->
trajectory design -> model frame -> fit -> compact results.

## Installation

You can install the development version of LEVAiM from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("Hugemiler/LEVAiM")
```

## Example

This example builds a small longitudinal dataset, models one microbial feature,
and compares time trajectories across treatment groups.

```{r example}
library(LEVAiM)

set.seed(1)

metadata <- expand.grid(
  subject = sprintf("S%02d", 1:12),
  week = c(0, 1, 2, 4, 8),
  KEEP.OUT.ATTRS = FALSE
)

metadata$sample_id <- sprintf("%s_W%s", metadata$subject, metadata$week)
metadata$treatment <- rep(
  rep(c("control", "fiber"), each = 6),
  each = length(unique(metadata$week))
)
metadata$subject <- factor(metadata$subject)
metadata$treatment <- factor(metadata$treatment)
rownames(metadata) <- metadata$sample_id

subject_shift <- rnorm(nlevels(metadata$subject), sd = 0.15)
names(subject_shift) <- levels(metadata$subject)

fiber_gain <- ifelse(
  metadata$treatment == "fiber",
  0.35 * log1p(metadata$week),
  0
)

assay <- data.frame(
  `Faecalibacterium prausnitzii` =
    2 +
    0.15 * metadata$week -
    0.015 * metadata$week^2 +
    fiber_gain +
    subject_shift[metadata$subject] +
    rnorm(nrow(metadata), sd = 0.12),
  check.names = FALSE
)
rownames(assay) <- rownames(metadata)

ds <- LongitudinalDataset(
  assays = list(taxa = assay),
  metadata = metadata,
  time_col = "week",
  subject_col = "subject"
)

spec <- model_spec(
  response = assay_feature("taxa", "Faecalibacterium prausnitzii"),
  subject = subject_var("subject"),
  time = time_var("week"),
  covariates = list(group_effect("treatment"))
)

traj <- trajectory(
  spec,
  design = varying_trajectory("treatment")
)

mf <- model_frame(
  ds,
  traj,
  control = trajectory_control(engine = "spline")
)

fit <- fit_trajectory(mf)
results <- trajectory_results(fit)

results
```

The returned `levaim_results` object keeps the original formula, fit metrics, and
smooth-term table so downstream code can inspect the fitted trajectory without
depending directly on the backend summary object.

```{r smooth-plot, fig.height = 4}
plot(fit)
```
