---
title: "00. The LEVAiM longitudinal microbiome course"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{00. The LEVAiM longitudinal microbiome course}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(LEVAiM)
```

# Start here

This is not a catalogue of unrelated functions. It is one analysis taught in
nine lectures.

We will ask a recurring biological question:

> How do infant gut microbial trajectories differ according to feeding practice
> around four months of age?

The Backhed profile gives us real longitudinal observations, repeated samples
from the same children, incomplete annotations, many sparse taxa, and a familiar
biological benchmark in early *Bifidobacterium* colonization. That makes it a
useful classroom dataset. The point of the course, however, is the operation:
how to move from files to an auditable longitudinal claim.

# The route through the course

| Lecture | Vignette | Question answered |
| --- | --- | --- |
| 1 | `data-loading` | What did we read, align, and retain? |
| 2 | `model-specification` | What exactly are we asking the model to estimate? |
| 3 | `feeding-4mo-trajectories` | Which taxa show feeding-associated trajectory differences? |
| 4 | `spline-fitting` | What does one fitted trajectory actually look like? |
| 5 | `validation` | Does the model predict held-out children sensibly? |
| 6 | `trajectory-analysis` | Which microbial trajectories move together or apart? |
| 7 | `response-families` | Are we modeling abundance, occurrence, or positive abundance? |
| 8 | `gp-crash-course` | What is a Gaussian process, and what assumptions does it make? |
| 9 | `gaussian-processes` | How do we fit and diagnose the real GP workflow? |

Begin with:

```{r start, eval = FALSE}
vignette("data-loading", package = "LEVAiM")
```

# One scientific contract

The lectures share four rules.

## Define exposure before fitting

`feeding_state` changes over time. Our primary exposure is the modal feeding
state observed from 3 through 5 months, called `feeding_4mo`. We derive it before
model-frame construction and retain its audit trail.

## Define the feature universe without reading the answers

The raw species table contains 2,011 columns after alignment. Most are nearly
empty: the median species occurs in less than 0.4% of samples. The course tests
species observed in at least 20% of aligned samples. In the bundled files this
produces 66 eligible species. The executable screen tests all 66; computation
is not used as a hidden second filter.

This is an unsupervised measurement-support rule. It does not inspect organism
names, feeding groups, effect sizes, or p-values. Every eligible taxon receives
the same model and enters the same multiplicity family.

## Keep subjects intact during validation

Repeated samples from one child are not independent train/test units. Every
validation split holds out complete subjects.

## Separate discovery from interpretation

We rank the complete eligible screen before attaching labels such as hallmark
colonizer or unexpected signal. A known organism may be used as a benchmark
afterward, but it does not receive privileged entry into the test universe.

# How the lectures run

Each lecture can be rendered on its own. It reconstructs the small amount of
state it needs rather than assuming that a previous vignette left objects in the
workspace.

The spline course is evaluated during ordinary vignette builds. The final GP
posterior is the exception because a diagnostics-grade Stan fit is deliberately
expensive. Lecture 9 evaluates parsing and compilation by default and gives one
explicit environment switch for the full posterior run. It also records the
diagnostics from the acceptance run; it does not disguise a short chain as
biological evidence.

# The analytical spine

Every lecture is a closer look at one part of the same pipeline:

```text
parse -> derive -> specify -> build -> compile -> fit
      -> validate -> predict -> interpret -> compare -> serialize
```

The important habit is not memorizing function names. It is knowing which
scientific decision is being made at each arrow.
