---
title: "BackhedF 2015 data loading for assay-wide trajectory discovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BackhedF 2015 data loading for assay-wide trajectory discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

# Objective

We start with a concrete question: can we organize a longitudinal infant gut
microbiome profile so that every taxonomic feature can be tested for a
feeding-associated trajectory?

This is not a generic import exercise. It is the first gate in the analysis.
If sample identifiers, metadata, and taxonomic profiles are not aligned exactly,
every downstream trajectory is suspect.

# Locate the real input files

The package includes a metadata table and a merged MetaPhlAn profile from the
BackhedF 2015 example files.

```{r paths}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}

metadata_path <- file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv")
taxa_path <- file.path(
  input_dir,
  "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
)

file.exists(metadata_path)
file.exists(taxa_path)
```

# Read metadata and taxonomic profiles

The metadata table carries the longitudinal design: sample, subject, age, and
feeding state. The merged MetaPhlAn table carries microbial relative abundance.
`read_metaphlan()` cleans the profile suffix so sample identifiers can be joined
directly.

```{r load-data}
metadata <- read.csv(
  metadata_path,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  taxa_path,
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

names(metadata)
dim(metadata)
dim(taxa)
head(metadata)
```

# Force sample alignment

This is the moment where the assay and metadata become one longitudinal
experiment. We keep only shared samples and impose identical row order.

```{r align-samples}
shared_samples <- intersect(metadata$sample, rownames(taxa))

alignment_audit <- data.frame(
  metadata_samples = length(metadata$sample),
  taxonomic_samples = nrow(taxa),
  shared_samples = length(shared_samples),
  metadata_only = length(setdiff(metadata$sample, rownames(taxa))),
  taxa_only = length(setdiff(rownames(taxa), metadata$sample))
)
alignment_audit

metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]

identical(rownames(taxa), rownames(metadata))
```

# Audit the full feature universe

Do not choose winners at import time. First inspect the full assay: how many
features are present, how sparse are they, and whether familiar early
colonizers are represented. Those familiar taxa are interpretation landmarks,
not filters.

```{r feature-universe}
feature_summary <- data.frame(
  feature = colnames(taxa),
  prevalence = colMeans(taxa > 0),
  mean_relative_abundance = colMeans(taxa),
  row.names = NULL
)

head(feature_summary[order(feature_summary$mean_relative_abundance, decreasing = TRUE), ])

colonizer_patterns <- c(
  "Bifidobacterium longum" = "s__Bifidobacterium_longum$",
  "Bifidobacterium breve" = "s__Bifidobacterium_breve$",
  "Bacteroides fragilis" = "s__Bacteroides_fragilis$",
  "Escherichia coli" = "s__Escherichia_coli$",
  "Enterococcus faecalis" = "s__Enterococcus_faecalis$",
  "Veillonella parvula" = "s__Veillonella_parvula$"
)

colonizer_features <- vapply(
  colonizer_patterns,
  function(pattern) grep(pattern, colnames(taxa), value = TRUE)[1],
  character(1)
)
colonizer_features <- colonizer_features[!is.na(colonizer_features)]

data.frame(
  colonizer = names(colonizer_features),
  mean_relative_abundance = colMeans(taxa[, colonizer_features, drop = FALSE]),
  row.names = NULL
)
```

# Build the longitudinal object

`LongitudinalDataset()` is the object we carry into every downstream vignette.

```{r dataset}
metadata$subject <- factor(metadata$subject)
metadata$feeding_state <- factor(metadata$feeding_state)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)

ds
```

# Continue

The next step is not fitting. It is declaring the biological question in a model
specification:

``` r
vignette("model-specification", package = "LEVAiM")
```
