---
title: "Loading data with LEVAiM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Baby steps with LEVAiM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Reading MetaPhlAn profiles

LEVAiM starts from bioBakery-style feature tables. The first supported input is MetaPhlAn 3/4 output.

```{r setup}
library(LEVAiM)
```

## Reading a merged MetaPhlAn table

Merged MetaPhlAn tables are usually produced with `merge_metaphlan_tables.py`.

```{r eval=FALSE}
taxa <- read_metaphlan(
  "some_merged_file.tsv",
  input_type = "merged",
  tax_level = "species"
)

dim(taxa)
head(taxa[, 1:5])
```

The returned object is a `data.frame` with samples as rows and taxa as columns.

## Reading multiple single-profile files

You can also pass a vector of single MetaPhlAn profile paths.

```{r eval=FALSE}
taxa <- read_metaphlan(
  c("s1.tsv", "s2.tsv"),
  input_type = "single",
  tax_level = "species"
)

dim(taxa)
head(taxa[, 1:5])
```

## Auto-detection

When possible, LEVAiM can try to detect whether the input is a single profile or a merged table.

```{r eval=FALSE}
taxa <- read_metaphlan(
  "some_merged_file.tsv",
  input_type = "auto"
)
```

## Keeping unclassified abundance

MetaPhlAn 4 outputs may include an `UNCLASSIFIED` percentage. By default, LEVAiM keeps it.

```{r eval=FALSE}
taxa <- read_metaphlan(
  "some_merged_file.tsv",
  input_type = "merged",
  include_unclassified = TRUE
)
```

To drop it:

```{r eval=FALSE}
taxa <- read_metaphlan(
  "some_merged_file.tsv",
  input_type = "merged",
  include_unclassified = FALSE
)
```

# Next step

The next LEVAiM milestone will combine feature tables with sample metadata into a longitudinal dataset object.
