---
title: "01. Meet and audit the longitudinal data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01. Meet and audit the longitudinal data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

**Course position:** Lecture 1 of 9<br>
**Syllabus:** `vignette("course-overview", package = "LEVAiM")`<br>
**Next:** `vignette("model-specification", package = "LEVAiM")`

# Today we earn the right to model

We start with a concrete question: can we organize a longitudinal infant gut
microbiome profile so that every taxonomic feature can be tested for a
feeding-associated trajectory?

This is not a generic import exercise. It is the first gate in the analysis.
If sample identifiers, metadata, and taxonomic profiles are not aligned exactly,
every downstream trajectory is suspect.

# Locate the real input files

The package includes a metadata table and a merged MetaPhlAn profile from the
BackhedF 2015 example files.

```{r paths}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}
if (!dir.exists(input_dir)) {
  input_dir <- system.file("extdata", package = "LEVAiM")
}

metadata_path <- file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv")
taxa_path <- file.path(
  input_dir,
  "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
)

file.exists(metadata_path)
file.exists(taxa_path)
```

# Read metadata and taxonomic profiles

The metadata table carries the longitudinal design: sample, subject, age, and
feeding state. The merged MetaPhlAn table carries microbial relative abundance.
`read_metaphlan()` cleans the profile suffix so sample identifiers can be joined
directly.

```{r load-data}
metadata <- read.csv(
  metadata_path,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  taxa_path,
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

names(metadata)
dim(metadata)
dim(taxa)
head(metadata)
```

# Force sample alignment

This is the moment where the assay and metadata become one longitudinal
experiment. We keep only shared samples and impose identical row order.

```{r align-samples}
shared_samples <- intersect(metadata$sample, rownames(taxa))

alignment_audit <- data.frame(
  metadata_samples = length(metadata$sample),
  taxonomic_samples = nrow(taxa),
  shared_samples = length(shared_samples),
  metadata_only = length(setdiff(metadata$sample, rownames(taxa))),
  taxa_only = length(setdiff(rownames(taxa), metadata$sample))
)
alignment_audit

metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]

identical(rownames(taxa), rownames(metadata))
```

# Audit the feature universe

Do not choose winners at import time. First inspect the full assay: how many
features are present, how sparse are they, and how many have enough measurement
support to enter the course screen.

```{r feature-universe}
feature_summary <- data.frame(
  feature = colnames(taxa),
  prevalence = colMeans(taxa > 0),
  mean_relative_abundance = colMeans(taxa),
  row.names = NULL
)

eligible_features <- feature_summary$feature[feature_summary$prevalence >= 0.20]

head(feature_summary[order(feature_summary$mean_relative_abundance, decreasing = TRUE), ])
length(eligible_features)
```

The 20% prevalence rule is declared without looking at organism names, feeding
groups, effect sizes, or p-values. The raw table remains available, but Lecture
3 will test 66 supported species rather than spending most of its effort on
1,945 sparsely observed columns. Lecture 3 executes the same model on every
eligible taxon.

# Build the longitudinal object

`LongitudinalDataset()` is the object we carry into every downstream vignette.

```{r dataset}
metadata$subject <- factor(metadata$subject)
metadata$feeding_state <- factor(metadata$feeding_state)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)

ds
```

# Pass the baton

We now know what was read, how rows were aligned, and which species have minimum
measurement support. The next step is not fitting. Lecture 2 declares what the
biological question means to the model.

``` r
vignette("model-specification", package = "LEVAiM")
```
