---
title: "BackhedF 2015 taxonomic trajectories"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BackhedF 2015 taxonomic trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

# Input files

This vignette uses the BackhedF_2015 files in `examples/input_files`:

```{r paths}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}

metadata_path <- file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv")
taxa_path <- file.path(
  input_dir,
  "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
)

file.exists(metadata_path)
file.exists(taxa_path)
```

# Load metadata and taxa

The metadata uses bare run accessions in `sample`. The merged MetaPhlAn table
uses the same accessions with a profile suffix, which `read_metaphlan()` removes
when `clean_sample_names = TRUE`.

```{r load-data}
metadata <- read.csv(
  metadata_path,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  taxa_path,
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

dim(metadata)
dim(taxa)
head(metadata)
taxa[1:5, 1:3]
```

# Align samples

The taxonomic profile contains additional runs that are not present in this
metadata file. For this first pipeline example, keep the shared samples and order
the assay table to match the metadata exactly.

```{r align-samples}
shared_samples <- intersect(metadata$sample, rownames(taxa))

length(shared_samples)
length(setdiff(metadata$sample, rownames(taxa)))
length(setdiff(rownames(taxa), metadata$sample))

metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample

taxa <- taxa[rownames(metadata), , drop = FALSE]

identical(rownames(taxa), rownames(metadata))
```

# Build a longitudinal dataset

The Backhed metadata already contains a subject identifier and infant age in
months. Feeding state is useful metadata, but it is missing for some later
samples, so this first trajectory uses a shared time smooth.

```{r dataset}
metadata$subject <- factor(metadata$subject)
metadata$feeding_state <- factor(
  ifelse(is.na(metadata$feeding_state), "Unknown", metadata$feeding_state)
)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)

ds
```

# Define and fit one trajectory

Choose the most abundant species in the aligned table and model its shared
trajectory over infant age.

```{r fit-trajectory}
feature <- names(sort(colMeans(taxa), decreasing = TRUE))[1]
feature

spec <- model_spec(
  response = assay_feature("taxa", feature),
  subject = subject_var("subject"),
  time = time_var("ageMonths")
)

traj <- trajectory(spec)

mf <- model_frame(
  ds,
  traj,
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 6
  )
)

mf

fit <- fit_trajectory(mf)
results <- trajectory_results(fit)

results
```

```{r plot-fit, fig.height = 4}
plot(fit)
```
