---
title: "Assay-wide discovery of feeding-associated trajectories"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assay-wide discovery of feeding-associated trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

# Overview

This vignette is the main discovery step. We fit the same feeding-stratified
trajectory model to every taxonomic feature, then inspect the result table for
both expected early-colonizer signals and unexpected organisms that deserve
follow-up.

# Prepare example data

```{r load-example}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

shared_samples <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]

metadata$subject <- factor(metadata$subject)
metadata$feeding_state <- factor(metadata$feeding_state)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)

ds_history <- derive_time_varying_features(
  ds,
  exposure = "feeding_state",
  summaries = list(
    feeding_4mo = exposure_window(3, 5, method = "modal"),
    ever_exclusive_breastfed = exposure_ever("ExclBreastFed"),
    feeding_transition = exposure_transition()
  )
)
```

# Test the full assay

Do not preselect bugs. The feature universe is the full taxonomic profile.

```{r select-features}
all_features <- colnames(taxa)
length(all_features)
```

# Screen current observed feeding state

`test_assay_features()` keeps the response as a single feature and repeats the
same specification across the selected assay columns.

This first screen asks whether trajectories differ by the feeding state observed
at each sample. That is a legitimate current-state question. It is not the same
as asking whether early feeding history predicts later trajectories.

The full screen is intentionally not evaluated during vignette builds because it
can fit thousands of models.

```{r current-feeding-screen, eval = FALSE}
current_feeding_tests <- test_assay_features(
  ds,
  assay = "taxa",
  features = all_features,
  subject = "subject",
  time = "ageMonths",
  covariates = list(group_effect("feeding_state", role = "time_varying")),
  design = varying_trajectory("feeding_state"),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 6,
    na_action = "explicit",
    sparse_min_n = 5
  )
)

current_feeding_tests$table
```

# Screen early feeding history

This second screen asks a different question: whether a subject-level derived
feeding-history feature, such as feeding around 4 months, modifies the
trajectory.

```{r early-feeding-screen, eval = FALSE}
early_feeding_tests <- test_assay_features(
  ds_history,
  assay = "taxa",
  features = all_features,
  subject = "subject",
  time = "ageMonths",
  covariates = list(
    group_effect("feeding_4mo", role = "derived_exposure"),
    group_effect("feeding_transition", role = "derived_exposure")
  ),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 6,
    na_action = "explicit",
    sparse_min_n = 5
  )
)

early_feeding_tests$table
```

# Rank the screen

After fitting, ranking happens in the result table. This is where known biology
and surprise findings enter the analysis.

```{r rank-results, eval = FALSE}
ranked_features <- current_feeding_tests$table[
  order(current_feeding_tests$table$p_adjust, current_feeding_tests$table$p_value),
]

head(ranked_features, 20)
```

# Label expected and unexpected signals

Known infant gut colonizers are useful landmarks, but they are not the only
possible discoveries. This annotation is deliberately post hoc: it labels the
screen, it does not define the screen.

```{r label-results, eval = FALSE}
known_colonizer_pattern <- paste(
  c(
    "Bifidobacterium",
    "Bacteroides",
    "Phocaeicola",
    "Escherichia",
    "Enterococcus",
    "Streptococcus",
    "Veillonella",
    "Lactobacillus"
  ),
  collapse = "|"
)

ranked_features$colonizer_label <- ifelse(
  grepl(known_colonizer_pattern, ranked_features$feature),
  "known_early_colonizer",
  "unexpected_signal"
)

ranked_features[
  ranked_features$status == "ok" & is.finite(ranked_features$p_adjust),
  c("feature", "p_value", "p_adjust", "test_term", "colonizer_label")
][1:20, ]
```

# Retain fits for downstream analysis

When trajectory comparisons are needed, retain fitted objects from the same
per-feature operation.

```{r retain-fits, eval = FALSE}
follow_up_features <- head(ranked_features$feature[ranked_features$status == "ok"], 8)

feature_tests_with_fits <- test_assay_features(
  ds,
  assay = "taxa",
  features = follow_up_features,
  subject = "subject",
  time = "ageMonths",
  covariates = list(group_effect("feeding_state", role = "time_varying")),
  design = varying_trajectory("feeding_state"),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 6,
    na_action = "explicit",
    sparse_min_n = 5
  ),
  keep_fits = TRUE
)

names(feature_tests_with_fits$fits)
```

# Continue

Use `vignette("trajectory-analysis", package = "LEVAiM")` to compare retained
feature trajectories and generate distance matrices.
