---
title: "Feeding around 4 months and infant gut trajectories"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Feeding around 4 months and infant gut trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

# Objective

This vignette asks a sharper question than "does the current feeding annotation
matter?"

> Are taxonomic trajectories associated with the feeding state recorded around
> 4 months of age?

The exposure is a derived subject-level feature: `feeding_4mo`, defined as the
modal `feeding_state` observed between 3 and 5 months. The response is still
feature-by-feature taxonomic abundance. The screen is not restricted to
favorite organisms. Hallmark early gut colonizers are used as interpretive
landmarks after the screen has run.

# Prepare the example data

```{r load-example}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

shared_samples <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]

metadata$subject <- factor(metadata$subject)
metadata$feeding_state <- factor(metadata$feeding_state)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)
```

# Derive feeding state around 4 months

`feeding_state` varies over time. For this question, using the raw column
directly would ask about the feeding state at each sample. Instead, derive the
history variable first, then model trajectories as a function of that derived
state.

```{r derive-feeding-4mo}
ds_4mo <- derive_time_varying_features(
  ds,
  exposure = "feeding_state",
  summaries = list(
    feeding_4mo = exposure_window(3, 5, method = "modal"),
    ever_exclusive_breastfed = exposure_ever("ExclBreastFed"),
    feeding_transition = exposure_transition()
  )
)

exposure_derivations(ds_4mo)
```

The resulting analysis strata are not balanced. That is part of the biological
and epidemiologic setting, not a nuisance to hide.

```{r feeding-4mo-distribution}
table(ds_4mo$metadata$feeding_4mo, useNA = "ifany")

subject_feeding <- unique(ds_4mo$metadata[, c("subject", "feeding_4mo")])
table(subject_feeding$feeding_4mo, useNA = "ifany")
```

In the completed run for this vignette, `feeding_4mo` was observed as
exclusive breastfeeding for 62 subjects, mixed feeding for 17 subjects, and
exclusive formula feeding for 10 subjects; 11 subjects lacked an observed
feeding state in the 3-to-5 month window. This is the window-centered exposure
needed for the 4-month visit, and it is much better populated than the earlier
2-to-4 month derivation.

# Screen the assay

The screen is intentionally assay-wide. For estimability and runtime, the
executed vignette run used taxa present in at least 5% of samples; this is a
prevalence filter, not a biological preselection. To fit every column in the
assay, replace `screen_features` with `all_features`.

```{r screen-features}
all_features <- colnames(taxa)
prevalence <- colMeans(taxa > 0, na.rm = TRUE)
screen_features <- names(prevalence[prevalence >= 0.05])

length(all_features)
length(screen_features)
```

The full screen is not evaluated during package vignette builds because it fits
one model per feature.

```{r feeding-4mo-screen, eval = FALSE}
feeding_4mo_tests <- test_assay_features(
  ds_4mo,
  assay = "taxa",
  features = screen_features,
  subject = "subject",
  time = "ageMonths",
  covariates = list(
    group_effect("feeding_4mo", role = "derived_exposure")
  ),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 6,
    na_action = "explicit",
    sparse_min_n = 5
  ),
  keep_fits = FALSE,
  error_action = "continue"
)

feeding_4mo_tests$table
```

# What the screen found

The completed 5%-prevalence screen fit 220 taxonomic features. All 220 fits
completed. The signal was broad:

```{r observed-fdr-summary}
observed_fdr_summary <- data.frame(
  threshold = c("q < 0.10", "q < 0.05", "q < 0.01"),
  significant_features = c(194, 181, 147),
  screened_features = 220
)

observed_fdr_summary
```

The strongest smooth terms were spread across the derived feeding groups:

```{r observed-term-summary}
observed_term_summary <- data.frame(
  term = c(
    "s(ageMonths):feeding_4moExclBreastFed",
    "s(ageMonths):feeding_4moExclFormulaFed",
    "s(ageMonths):feeding_4moMixed",
    "s(ageMonths):feeding_4moUnknown"
  ),
  significant_features_q05 = c(115, 20, 36, 10)
)

observed_term_summary
```

This is not a tidy result in which only the obvious infant organisms light up.
That is the point of the wrapper. The screen is allowed to find broad
maturation structure, missingness-linked structure, and unexpected taxa.

# Hallmark early colonizers enter after ranking

The screen itself does not preselect taxa. After fitting, label familiar infant
gut colonizers so the result table can be read biologically.

```{r posthoc-labeling, eval = FALSE}
known_colonizer_pattern <- paste(
  c(
    "Bifidobacterium",
    "Bacteroides",
    "Phocaeicola",
    "Escherichia",
    "Enterococcus",
    "Streptococcus",
    "Veillonella",
    "Lactobacillus",
    "Akkermansia"
  ),
  collapse = "|"
)

ranked_features <- feeding_4mo_tests$table[
  order(feeding_4mo_tests$table$p_adjust, feeding_4mo_tests$table$p_value),
]

ranked_features$colonizer_label <- ifelse(
  grepl(known_colonizer_pattern, ranked_features$feature),
  "known_landmark",
  "unexpected_or_less_canonical"
)

table(
  ranked_features$colonizer_label[
    ranked_features$status == "ok" & ranked_features$p_adjust < 0.05
  ]
)
```

In the completed run, 39 significant rows matched the hallmark-colonizer label
and 142 significant rows did not. That means the expected organisms are present,
but they are not the whole story.

The leading results included familiar infant-associated taxa and organisms more
suggestive of broader gut maturation:

```{r observed-top-hits}
observed_top_hits <- data.frame(
  feature = c(
    "Roseburia_faecis",
    "Bacteroides_uniformis",
    "Eubacterium_rectale",
    "Faecalibacterium_prausnitzii",
    "Bacteroides_ovatus",
    "Blautia_wexlerae",
    "Clostridium_sp_AF34_10BH",
    "Lachnospira_eligens",
    "Dialister_invisus",
    "Ruminococcus_bromii",
    "Fusicatenibacter_saccharivorans",
    "Anaerostipes_hadrus",
    "Bifidobacterium_longum"
  ),
  leading_term = rep("ExclBreastFed", 13),
  q_value = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0),
  deviance_explained = c(
    0.226,
    0.527,
    0.282,
    0.446,
    0.484,
    0.393,
    0.167,
    0.228,
    0.183,
    0.238,
    0.289,
    0.483,
    0.411
  )
)

observed_top_hits
```

`Bacteroides_uniformis`, `Bacteroides_ovatus`, and
`Bifidobacterium_longum` are reassuring landmarks. But the top of the table is
also rich in anaerobes and short-chain-fatty-acid-associated organisms. The
working biological reading is that feeding state around 4 months is associated
with the tempo and shape of community maturation, not merely with the presence
or absence of one iconic infant taxon.

# Retain fits for trajectory comparisons

Once the screen has ranked features, refit a small follow-up set with retained
fits. This is the bridge from "which features are significant?" to "which
trajectories look coordinated?"

```{r retained-fits, eval = FALSE}
follow_up_features <- head(
  ranked_features$feature[ranked_features$status == "ok"],
  8
)

feeding_4mo_followup <- test_assay_features(
  ds_4mo,
  assay = "taxa",
  features = follow_up_features,
  subject = "subject",
  time = "ageMonths",
  covariates = list(
    group_effect("feeding_4mo", role = "derived_exposure")
  ),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 6,
    na_action = "explicit",
    sparse_min_n = 5
  ),
  keep_fits = TRUE,
  error_action = "continue"
)
```

The retained follow-up run used the first eight ranked taxa. The fitted
features were:

```{r observed-followup-features}
observed_followup_features <- data.frame(
  feature = c(
    "Roseburia_faecis",
    "Bacteroides_uniformis",
    "Eubacterium_rectale",
    "Faecalibacterium_prausnitzii",
    "Bacteroides_ovatus",
    "Blautia_wexlerae",
    "Clostridium_sp_AF34_10BH",
    "Lachnospira_eligens"
  ),
  leading_term = rep("ExclBreastFed", 8),
  deviance_explained = c(0.226, 0.527, 0.282, 0.446, 0.484, 0.393, 0.167, 0.228)
)

observed_followup_features
```

# Compare trajectories

`compare_trajectories()` returns the analysis layer biologists and
biostatisticians will ask for: a long data frame. Each row is a feature pair,
a metadata stratum, a time window, a metric, and its descriptive evidence.

```{r compare-trajectories, eval = FALSE}
trajectory_comparison <- compare_trajectories(
  feeding_4mo_followup,
  n = 30,
  metrics = c(
    "fitted_correlation",
    "fitted_distance",
    "derivative_correlation"
  )
)

head(trajectory_comparison)
```

In the completed follow-up, the comparison table had 336 rows across the
exclusive-breastfed, exclusive-formula-fed, mixed, and unknown `feeding_4mo`
groups. Fitted correlations were often strongly positive, but not uniformly
so. The smallest fitted distances occurred for feature pairs with nearly
identical fitted shapes inside a feeding-history stratum, such as
`Bacteroides_uniformis` with `Bacteroides_ovatus` in the unknown stratum and
`Bacteroides_ovatus` with `Blautia_wexlerae` in the formula-fed stratum. The
first-eight follow-up set therefore reads as a coordinated maturation block
rather than a collection of unrelated discoveries.

The p-values in this table are deliberately labeled as descriptive. The fitted
grid points are not independent biological observations.

# Generate a distance matrix

Heatmap inputs are derived from the long comparison table. This keeps the
statistical evidence table as the source of truth and lets plotting code choose
which group and time window to visualize.

```{r distance-matrix, eval = FALSE}
feeding_4mo_distance <- trajectory_distance_matrix(
  trajectory_comparison
)

feeding_4mo_distance
```

The completed run produced an 8-by-8 fitted-distance matrix for the first
available fitted-distance stratum, the exclusive-breastfed trajectory group.
The matrix is heatmap-ready, but its biological reading still depends on the
chosen stratum and the stability of that stratum. The long table remains the
source of truth when a specific feeding group or time window is required.

# Interpretation

This vignette supports three conclusions.

First, deriving `feeding_4mo` changes the question in a useful way. The model is
no longer asking only whether the sample's current feeding state matches the
sample's current abundance. It asks whether early feeding history stratifies the
trajectory.

Second, the result is stronger and broader than a hallmark-only story.
Canonical infant-associated taxa appear, but the strongest results also include
adult-like anaerobes and community-maturation organisms. That is biologically
plausible: feeding practice around 4 months can mark a developmental regime,
not just a single substrate at a single visit.

Third, the corrected 3-to-5 month window materially improves exposure support.
Unknown `feeding_4mo` is no longer the dominant stratum, although exclusive
formula feeding remains smaller than exclusive breastfeeding. Those levels
should remain in the model frame so samples are not silently discarded, but
formula-specific claims should still be treated as follow-up hypotheses.
