---
title: "03. Discover feeding-associated trajectories"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{03. Discover feeding-associated trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(LEVAiM)
```

**Course position:** Lecture 3 of 9<br>
**Previously:** `vignette("model-specification", package = "LEVAiM")`<br>
**Next:** `vignette("spline-fitting", package = "LEVAiM")`

# Today we run the screen

In Lecture 2 we translated the biological question into a model specification.
Now every taxon with adequate measurement support gets the same chance to
answer it:

> Between 3 and 5 months, do fitted abundance trajectories differ among infants
> classified as exclusively breastfed, exclusively formula fed, or mixed fed?

We will not begin with a list of favorite organisms. We will define eligibility
using prevalence alone, test every eligible species, correct that complete set
of p-values, and only then ask whether familiar or surprising taxa appear.

# Reconstruct the course dataset

Each lecture runs on its own, so we repeat the short parsing block rather than
depending on objects left behind by another vignette.

```{r load-course-data}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) input_dir <- file.path("examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- system.file("extdata", package = "LEVAiM")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

shared_samples <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]
metadata$subject <- factor(metadata$subject)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)
ds <- derive_time_varying_features(
  ds,
  exposure = "feeding_state",
  summaries = list(feeding_4mo = exposure_window(3, 5, method = "modal"))
)
```

# Define who is compared

Before touching the taxa, inspect the derived exposure. Sample counts and
subject counts answer different questions; repeated samples make the first look
larger.

```{r exposure-support}
table(ds$metadata$feeding_4mo, useNA = "ifany")

subject_feeding <- unique(ds$metadata[c("subject", "feeding_4mo")])
table(subject_feeding$feeding_4mo, useNA = "ifany")
```

Unknown feeding histories remain visible in the model frame, but they are not a
fourth biological feeding practice. The test below names the three levels in
the hypothesis explicitly.

# Reduce 2,011 sparse columns to an estimable universe

The raw profile contains thousands of species because taxonomic profilers can
name organisms seen in very few samples. Fitting all of them is not
comprehensive; it is mostly testing absence.

```{r eligible-universe}
prevalence <- colMeans(taxa > 0, na.rm = TRUE)
eligible_features <- names(prevalence[prevalence >= 0.20])

data.frame(
  raw_species = ncol(taxa),
  eligible_species = length(eligible_features),
  prevalence_threshold = 0.20
)
```

The 20% threshold is declared before fitting and ignores taxon identity,
feeding group, effect size, and outcome significance. In these files it reduces
2,011 species to 66. We test all 66. A teaching budget is not a scientific
reason to remove eligible outcomes from the multiplicity family.

# Fit trajectories first; localize the contrast second

Every model uses every informative longitudinal observation. The 3–5 month
window does **not** subset the model to a cross-sectional test. It defines the
domain over which we integrate the uncentered contrast between smooth fitted
trajectories. A feeding-associated level difference around four months is
therefore signal, not a nuisance offset to be subtracted away.

These children were sampled at three biological waves. Slightly different
recorded ages do not create six independent temporal degrees of freedom, so we
set `spline_k = 3`.

```{r assay-screen}
fast_null <- trajectory_null_control(
  method = "gaussian_approximation",
  simulations = 4999,
  seed = 11
)

feeding_screen <- test_assay_features(
  ds,
  assay = "taxa",
  features = eligible_features,
  subject = "subject",
  time = "ageMonths",
  covariates = list(
    group_effect("feeding_4mo", role = "derived_exposure")
  ),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 3,
    na_action = "explicit",
    sparse_min_n = 5
  ),
  hypothesis = "critical_period_level",
  test_covariate = "feeding_4mo",
  test_levels = c("ExclBreastFed", "ExclFormulaFed", "Mixed"),
  test_window = c(3, 5),
  null_control = fast_null,
  keep_fits = FALSE,
  error_action = "continue"
)
```

This is the middle layer between one hand-built model and an opaque black-box
pipeline. The result is a data frame with one row per eligible species, one
declared smooth-trajectory hypothesis per row, and one multiplicity correction
across all 66 rows.

This first pass uses a conditional Gaussian coefficient approximation. It is
fast enough to audit the complete feature universe and precise enough to rank
small p-values, but it is not an empirical repeated-measures null. Its q-values
are provisional until the design-specific plasmode and subject-cluster
bootstrap gates agree.

# Read the table before telling a story

```{r screen-summary}
table(feeding_screen$table$status)
feeding_screen$fdr_resolution
sum(feeding_screen$table$p_adjust < 0.05, na.rm = TRUE)
sum(feeding_screen$table$p_adjust < 0.10, na.rm = TRUE)

ranked <- feeding_screen$table[
  order(feeding_screen$table$p_adjust, feeding_screen$table$p_value),
]
head(ranked[c(
  "feature", "p_value", "p_adjust", "model_effect", "model_std_error",
  "effect_level_1", "effect_level_2", "n", "min_group_n", "null_method"
)], 10)
```

`model_effect` is the strongest pair's average fitted difference over months
3–5 on the `log1p` model scale. It is computed from the smooth trajectories,
not from group proportions in a window. The raw window summaries remain in
`descriptive_summary` so we can catch a model result that contradicts its own
observations.

# Ask whether zeros carry a different answer

Relative-abundance zeros can encode absence, undersampling, or both. One
Gaussian curve cannot tell us whether feeding changes colonization probability
or abundance among colonized infants. We therefore fit two more longitudinal
trajectory screens over the same 66-feature family.

```{r component-screens}
feeding_occurrence <- test_assay_features(
  ds, assay = "taxa", features = eligible_features,
  subject = "subject", time = "ageMonths",
  covariates = list(group_effect("feeding_4mo", role = "derived_exposure")),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(spline_k = 3, na_action = "explicit", sparse_min_n = 5),
  hypothesis = "occurrence_difference",
  test_covariate = "feeding_4mo",
  test_levels = c("ExclBreastFed", "ExclFormulaFed", "Mixed"),
  test_window = c(3, 5), null_control = fast_null,
  error_action = "continue"
)

feeding_positive <- test_assay_features(
  ds, assay = "taxa", features = eligible_features,
  subject = "subject", time = "ageMonths",
  covariates = list(group_effect("feeding_4mo", role = "derived_exposure")),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(transform = "log1p", spline_k = 3,
    na_action = "explicit", sparse_min_n = 5),
  hypothesis = "positive_abundance_difference",
  test_covariate = "feeding_4mo",
  test_levels = c("ExclBreastFed", "ExclFormulaFed", "Mixed"),
  test_window = c(3, 5), null_control = fast_null,
  error_action = "continue"
)

data.frame(
  estimand = c("total abundance", "occurrence", "positive abundance"),
  successful = c(
    sum(feeding_screen$table$status == "ok"),
    sum(feeding_occurrence$table$status == "ok"),
    sum(feeding_positive$table$status == "ok")
  ),
  provisional_q_below_0.10 = c(
    sum(feeding_screen$table$p_adjust < 0.10, na.rm = TRUE),
    sum(feeding_occurrence$table$p_adjust < 0.10, na.rm = TRUE),
    sum(feeding_positive$table$p_adjust < 0.10, na.rm = TRUE)
  )
)

estimand_audit <- do.call(rbind, list(
  transform(feeding_screen$table, estimand = "total_abundance"),
  transform(feeding_occurrence$table, estimand = "occurrence"),
  transform(feeding_positive$table, estimand = "positive_abundance")
))
rownames(estimand_audit) <- NULL

head(estimand_audit[c(
  "estimand", "feature", "status", "model_effect", "model_std_error",
  "p_value", "p_adjust", "n", "n_detected", "detection_rate"
)])
```

Four positive-abundance models fail because at least one feeding group has no
positive observations for that species. That is a support failure worth
reporting, not a zero to conceal. The long table is convenient for comparison,
but each estimand keeps its own BH family because the three rows for one species
answer different biological questions.

# Stress the result across composition and rank

```{r sensitivity-screens}
feeding_renormalized <- test_assay_features(
  ds, assay = "taxa", features = eligible_features,
  subject = "subject", time = "ageMonths",
  covariates = list(group_effect("feeding_4mo", role = "derived_exposure")),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(transform = "log1p", spline_k = 3,
    na_action = "explicit", sparse_min_n = 5),
  hypothesis = "critical_period_level", test_covariate = "feeding_4mo",
  test_levels = c("ExclBreastFed", "ExclFormulaFed", "Mixed"),
  test_window = c(3, 5), null_control = fast_null,
  composition_policy = "renormalize", error_action = "continue"
)

genus_taxa <- aggregate_taxa(taxa, tax_level = "genus")
genus_ds <- ds
genus_ds$assays$taxa <- genus_taxa
eligible_genera <- names(which(colMeans(genus_taxa > 0) >= 0.20))

feeding_genera <- test_assay_features(
  genus_ds, assay = "taxa", features = eligible_genera,
  subject = "subject", time = "ageMonths",
  covariates = list(group_effect("feeding_4mo", role = "derived_exposure")),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(transform = "log1p", spline_k = 3,
    na_action = "explicit", sparse_min_n = 5),
  hypothesis = "critical_period_level", test_covariate = "feeding_4mo",
  test_levels = c("ExclBreastFed", "ExclFormulaFed", "Mixed"),
  test_window = c(3, 5), null_control = fast_null,
  error_action = "continue"
)

community_ds <- ds
community_ds$assays$community <- community_features(taxa)
feeding_community <- test_assay_features(
  community_ds, assay = "community",
  subject = "subject", time = "ageMonths",
  covariates = list(group_effect("feeding_4mo", role = "derived_exposure")),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(spline_k = 3, na_action = "explicit", sparse_min_n = 5),
  hypothesis = "critical_period_level", test_covariate = "feeding_4mo",
  test_levels = c("ExclBreastFed", "ExclFormulaFed", "Mixed"),
  test_window = c(3, 5), null_control = fast_null,
  error_action = "continue"
)

stats::cor(
  feeding_screen$table$p_value,
  feeding_renormalized$table$p_value,
  method = "spearman",
  use = "complete.obs"
)
sum(feeding_genera$table$p_adjust < 0.10, na.rm = TRUE)
feeding_community$table[c(
  "feature", "model_effect", "effect_level_1", "effect_level_2",
  "p_value", "p_adjust"
)]
```

The supplied and renormalized species rankings are nearly identical in this
run, while genus aggregation changes which biological units lead. The
community table asks a broader succession question without pretending that one
species must carry the full feeding signal. In the fast audit, breastfed infants
have lower fitted richness and Shannon diversity than formula-fed infants near
four months, consistent with delayed acquisition of an adult-like diverse
community. These remain provisional coefficient-null results.

That is the right use of sensitivity analysis: it shows which conclusions
survive a reasonable representation change without pretending species, genera,
and community summaries are interchangeable hypotheses.

# Put the community conclusion through the empirical gate

Five prespecified community endpoints are small enough to support empirical
refitting in the executable lecture. The reduced null removes feeding group but
retains the shared smooth trajectory and subject structure. One wild-bootstrap
multiplier moves all residuals from a child together.

```{r community-empirical-null}
feeding_community_empirical <- test_assay_features(
  community_ds, assay = "community",
  subject = "subject", time = "ageMonths",
  covariates = list(group_effect("feeding_4mo", role = "derived_exposure")),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(spline_k = 3, na_action = "explicit", sparse_min_n = 5),
  hypothesis = "critical_period_level", test_covariate = "feeding_4mo",
  test_levels = c("ExclBreastFed", "ExclFormulaFed", "Mixed"),
  test_window = c(3, 5),
  null_control = trajectory_null_control(
    method = "cluster_wild_bootstrap",
    simulations = 199,
    seed = 29,
    cores = 4
  ),
  error_action = "stop"
)

feeding_community_empirical$table[c(
  "feature", "model_effect", "effect_level_1", "effect_level_2",
  "p_value", "p_adjust", "null_method", "null_failed_simulations"
)]
```

Richness and Shannon diversity are lower in the exclusively breastfed
trajectory than in the exclusively formula-fed trajectory around four months
(`p = 0.020` and `0.025`; both `q = 0.0625`). Dominance points in the reciprocal
direction (`p = 0.075`, `q = 0.125`). This is a recognizable early-life
succession result: exclusive breastfeeding is associated with a less rich,
less even, more dominated community at this critical period, consistent with
later acquisition of an adult-like diverse community.

The species screen remains useful for localization, but the broad biological
claim is validated at the community level where it was actually formulated.

As an independent design-null audit, subject-label permutation preserves every
child's complete trajectory and reassigns only the subject-static feeding label.
It gives `p = 0.025` for richness and `p = 0.045` for Shannon diversity; both
adjust to `q = 0.1125` across the five community endpoints. The direction and
raw evidence therefore reproduce across null constructions, while the
`q < 0.10` classification does not. We report a supported community succession
signal whose multiplicity boundary remains calibration-sensitive.

# Now audit hallmark colonizers

Only after ranking do we attach biological labels.

```{r hallmark-audit}
hallmark_pattern <- paste(
  c(
    "Bifidobacterium", "Bacteroides", "Escherichia", "Enterococcus",
    "Staphylococcus", "Streptococcus", "Veillonella", "Clostridium"
  ),
  collapse = "|"
)

hallmark_rows <- grepl(hallmark_pattern, ranked$feature)
ranked[hallmark_rows, c(
  "feature", "p_value", "p_adjust", "model_effect",
  "effect_level_1", "effect_level_2"
)][1:min(20, sum(hallmark_rows)), ]
```

This ordering matters. *Bifidobacterium* is a valuable positive-control family,
but it did not receive a private screen. Known biology is used to judge whether
the resulting evidence is plausible, not to decide which columns were allowed
to generate evidence.

Under the `k = 3` total-abundance estimand, *Bifidobacterium breve* is among the
provisional findings and the fitted breast-versus-formula contrast has the
expected direction. *B. longum* has the same direction but weaker evidence.
That is a biologically recognizable result, not permission to stop validating
the null.

# What leaves this lecture

We now have:

- a reproducible all-eligible 66-species screen;
- a ranked table with raw and FDR-adjusted evidence;
- an integrated critical-period contrast of full smooth trajectories;
- separate occurrence and positive-abundance trajectory screens;
- explicit composition and genus-rank sensitivity analyses;
- an explicitly provisional fast-null tier awaiting empirical confirmation;
- a post-ranking audit of hallmark colonizers.

Lecture 4 takes one benchmark from this screen and asks a harder question:
what is the fitted model actually saying about the observations, uncertainty,
and rate of change?

```{r next-lecture, eval = FALSE}
vignette("spline-fitting", package = "LEVAiM")
```
