---
title: "Feeding around 4 months and infant gut trajectories"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Feeding around 4 months and infant gut trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

# Objective

This vignette asks a sharper question than "does the current feeding annotation
matter?"

> Are taxonomic trajectories associated with the feeding state recorded around
> 4 months of age?

The exposure is a derived subject-level feature: `feeding_4mo`, defined as the
modal `feeding_state` observed between 3 and 5 months. The response is still
feature-by-feature taxonomic abundance. The screen is not restricted to
favorite organisms. Hallmark early gut colonizers are used as interpretive
landmarks after the screen has run.

# Prepare the example data

```{r load-example}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

shared_samples <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]

metadata$subject <- factor(metadata$subject)
metadata$feeding_state <- factor(metadata$feeding_state)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)
```

# Derive feeding state around 4 months

`feeding_state` varies over time. For this question, using the raw column
directly would ask about the feeding state at each sample. Instead, derive the
history variable first, then model trajectories as a function of that derived
state.

```{r derive-feeding-4mo}
ds_4mo <- derive_time_varying_features(
  ds,
  exposure = "feeding_state",
  summaries = list(
    feeding_4mo = exposure_window(3, 5, method = "modal"),
    ever_exclusive_breastfed = exposure_ever("ExclBreastFed"),
    feeding_transition = exposure_transition()
  )
)

exposure_derivations(ds_4mo)
```

The resulting analysis strata are not balanced. That is part of the biological
and epidemiologic setting, not a nuisance to hide.

```{r feeding-4mo-distribution}
table(ds_4mo$metadata$feeding_4mo, useNA = "ifany")

subject_feeding <- unique(ds_4mo$metadata[, c("subject", "feeding_4mo")])
table(subject_feeding$feeding_4mo, useNA = "ifany")
```

In the completed run for this vignette, `feeding_4mo` was observed as
exclusive breastfeeding for 62 subjects, mixed feeding for 17 subjects, and
exclusive formula feeding for 10 subjects; 11 subjects lacked an observed
feeding state in the 3-to-5 month window. This is the window-centered exposure
needed for the 4-month visit, and it is much better populated than the earlier
2-to-4 month derivation.

# Screen the assay

The screen is intentionally assay-wide. For estimability and runtime, the
executed vignette run used taxa present in at least 5% of samples; this is a
prevalence filter, not a biological preselection. To fit every column in the
assay, replace `screen_features` with `all_features`.

```{r screen-features}
all_features <- colnames(taxa)
prevalence <- colMeans(taxa > 0, na.rm = TRUE)
screen_features <- names(prevalence[prevalence >= 0.05])

length(all_features)
length(screen_features)
```

The full screen is not evaluated during package vignette builds because it fits
one group-specific trajectory model per feature and calibrates a maximum
standardized contrast statistic from that model's coefficient covariance. The
fit uses all observed ages, but inference is prespecified over 3 to 5 months,
the same biological window used to construct `feeding_4mo`. Missing-exposure
samples remain in the fit while only the three observed feeding states enter
the biological test.

```{r feeding-4mo-screen, eval = FALSE}
feeding_4mo_tests <- test_assay_features(
  ds_4mo,
  assay = "taxa",
  features = screen_features,
  subject = "subject",
  time = "ageMonths",
  covariates = list(
    group_effect("feeding_4mo", role = "derived_exposure")
  ),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 6,
    na_action = "explicit",
    sparse_min_n = 5
  ),
  hypothesis = "trajectory_difference",
  test_covariate = "feeding_4mo",
  test_levels = c("ExclBreastFed", "ExclFormulaFed", "Mixed"),
  test_time_values = seq(3, 5, length.out = 30),
  test_simulations = 1999,
  test_seed = 1,
  keep_fits = FALSE,
  error_action = "continue"
)

feeding_4mo_tests$table
```

# What the screen found

The completed 5%-prevalence screen tested 220 taxonomic features. All 220
completed. The localized omnibus test identified a selective feeding-window
signal:

```{r observed-fdr-summary}
observed_fdr_summary <- data.frame(
  threshold = c("q < 0.10", "q < 0.05", "q < 0.01"),
  significant_features = c(17, 12, 0),
  screened_features = 220
)

observed_fdr_summary
```

Every row now carries one hypothesis and one null, rather than whichever
group-specific smooth happened to have the smallest p-value:

```{r observed-term-summary}
observed_test_contract <- data.frame(
  field = c("hypothesis", "null", "method", "multiplicity"),
  value = c(
    "trajectory_difference",
    "equal fitted trajectories across observed feeding_4mo groups from 3 to 5 months",
    "maximum standardized contrast calibrated from coefficient covariance",
    "Benjamini-Hochberg across 220 features"
  )
)

observed_test_contract
```

The previous 181-feature result was not evidence of a near-global feeding
effect. It selected the smallest p-value among several smooth terms. The later
nested-GAM replacement was also unsuitable: its difference smooths could be
shrunk to zero and its conclusions changed with the spline basis. The current
test asks directly whether any fitted group contrast separates inside the
prespecified feeding window while accounting for correlation across the time
grid.

# Hallmark early colonizers enter after ranking

The screen itself does not preselect taxa. After fitting, label familiar infant
gut colonizers so the result table can be read biologically.

```{r posthoc-labeling, eval = FALSE}
known_colonizer_pattern <- paste(
  c(
    "Bifidobacterium",
    "Bacteroides",
    "Phocaeicola",
    "Escherichia",
    "Enterococcus",
    "Streptococcus",
    "Veillonella",
    "Lactobacillus",
    "Akkermansia"
  ),
  collapse = "|"
)

ranked_features <- feeding_4mo_tests$table[
  order(feeding_4mo_tests$table$p_adjust, feeding_4mo_tests$table$p_value),
]

ranked_features$colonizer_label <- ifelse(
  grepl(known_colonizer_pattern, ranked_features$feature),
  "known_landmark",
  "unexpected_or_less_canonical"
)

table(
  ranked_features$colonizer_label[
    ranked_features$status == "ok" & ranked_features$p_adjust < 0.05
  ]
)
```

The strict q < 0.05 set contains 12 features. Familiar early-life organisms are
present, but the result is not restricted to a hand-picked colonizer list.

The leading results included familiar infant-associated taxa and organisms more
suggestive of broader gut maturation:

```{r observed-top-hits}
observed_top_hits <- data.frame(
  feature = c(
    "Sellimonas_intestinalis",
    "Flavonifractor_plautii",
    "Granulicatella_SGB8255",
    "Lacticaseibacillus_rhamnosus",
    "Actinomyces_SGB17168",
    "Enterococcus_casseliflavus",
    "Actinomyces_sp_HMSC065F12",
    "TM7_phylum_sp_oral_taxon_352",
    "Robinsoniella_peoriensis",
    "Actinomyces_sp_ph3",
    "Citrobacter_freundii",
    "Streptococcus_lactarius"
  ),
  p_value = c(rep(0.0005, 8), 0.001, 0.001, 0.0015, 0.0025),
  q_value = c(rep(0.01375, 8), 0.022, 0.022, 0.030, 0.04583),
  max_separation_month = c(
    4.17, 4.38, 4.24, 4.66, 4.03, 4.10, 4.45, 4.03,
    4.38, 3.83, 3.00, 4.52
  )
)

observed_top_hits
```

The strongest separations cluster near four months, as they should for this
question. `Enterococcus`, `Streptococcus`, `Lacticaseibacillus`, and
`Actinomyces` provide recognizable early-life signals. These remain statistical
discoveries rather than automatic mechanistic claims.

# Audit the Bifidobacterium positive control

Feeding-associated bifidobacteria are a required biological audit, not a reason
to preselect the screen. The complete genus pattern is therefore inspected
after ranking:

```{r observed-bifidobacteria}
observed_bifidobacteria <- data.frame(
  feature = c(
    "Bifidobacterium_breve",
    "Bifidobacterium_longum",
    "Bifidobacterium_adolescentis",
    "Bifidobacterium_pseudocatenulatum",
    "Bifidobacterium_catenulatum",
    "Bifidobacterium_dentium",
    "Bifidobacterium_bifidum"
  ),
  p_value = c(0.0055, 0.0855, 0.0865, 0.359, 0.400, 0.423, 0.595),
  q_value = c(0.0756, 0.4325, 0.4325, 1, 1, 1, 1),
  max_separation_month = c(4.79, 4.52, 3.62, 5.00, 3.69, 4.17, 5.00)
)

observed_bifidobacteria
```

`B. breve` is suggestive after assay-wide FDR correction (q = 0.0756).
`B. longum` has the expected localized ordering but weaker three-group omnibus
evidence (p = 0.0855). In the prespecified exclusive-breastfed versus
exclusive-formula contrast over 3 to 5 months, its covariance-calibrated p-value
is 0.0355 and its maximum separation occurs near 4.52 months. The fitted
exclusive-breastfed minus formula difference near four months is approximately
0.77 log units. Mixed feeding is intermediate, which weakens the three-group
test without erasing the targeted contrast.

# Summarize trajectory moments

Pointwise contrasts locate separation, but biologists also need a compact
description of what each curve does. `trajectory_moments()` converts fitted
curves into a long table of level, AUC, temporal variation, amplitude, timing,
trend, velocity, and conditional turning-point summaries. The model is still
fit over all ages; moments are reported inside declared biological windows.

```{r trajectory-moments, eval = FALSE}
moment_windows <- data.frame(
  label = c("early", "feeding_window", "later"),
  start = c(0, 3, 5),
  end = c(3, 5, 12)
)

b_longum_feature <- grep(
  "s__Bifidobacterium_longum$",
  colnames(taxa),
  value = TRUE
)

b_longum_followup <- test_assay_features(
  ds_4mo,
  assay = "taxa",
  features = b_longum_feature,
  subject = "subject",
  time = "ageMonths",
  covariates = list(
    group_effect("feeding_4mo", role = "derived_exposure")
  ),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 6,
    na_action = "explicit",
    sparse_min_n = 5
  ),
  hypothesis = "trajectory_difference",
  test_covariate = "feeding_4mo",
  test_levels = c("ExclBreastFed", "ExclFormulaFed", "Mixed"),
  test_time_values = seq(3, 5, length.out = 30),
  keep_fits = TRUE,
  error_action = "stop"
)

b_longum_moments <- trajectory_moments(
  b_longum_followup$fits[[b_longum_feature]],
  n = 181,
  windows = moment_windows,
  levels = c("ExclBreastFed", "ExclFormulaFed", "Mixed"),
  draws = 500,
  seed = 7
)

b_longum_moment_comparisons <- compare_trajectory_moments(
  b_longum_moments,
  reference = "ExclFormulaFed"
)
```

The same fitted objects and DataFrames drive the plotting layer:

```{r moment-plots, eval = FALSE}
plot_trajectory_report(
  b_longum_followup$fits[[b_longum_feature]],
  levels = c("ExclBreastFed", "ExclFormulaFed", "Mixed"),
  window = c(3, 5),
  points = TRUE
)

b_longum_contrasts <- trajectory_contrasts(
  b_longum_followup$fits[[b_longum_feature]],
  time_values = seq(3, 5, length.out = 80),
  levels = c("ExclBreastFed", "ExclFormulaFed", "Mixed")
)

plot_trajectory_contrasts(b_longum_contrasts)

plot_trajectory_moments(
  b_longum_moments,
  moment = c("mean_level", "auc", "linear_trend", "peak_time"),
  window = "feeding_window"
)

plot_trajectory_moment_comparisons(
  b_longum_moment_comparisons,
  moment = c("mean_level", "auc", "linear_trend", "peak_value"),
  window = "feeding_window"
)
```

The completed `B. longum` feeding-window report is compact enough to inspect
directly:

```{r observed-b-longum-moments}
observed_b_longum_moments <- data.frame(
  feeding_group = c("ExclBreastFed", "ExclFormulaFed", "Mixed"),
  mean_level = c(2.160, 1.409, 1.673),
  mean_level_lower = c(1.900, 0.779, 1.170),
  mean_level_upper = c(2.421, 1.997, 2.126),
  auc_3_to_5 = c(4.184, 2.730, 3.241),
  linear_trend = c(0.191, 0.016, 0.103),
  peak_time = c(4.991, 4.321, 4.991),
  peak_status = c(
    "boundary_only",
    "interior_supported",
    "boundary_only"
  )
)

observed_b_longum_moments
```

Exclusive breastfeeding exceeds formula feeding by 0.751 fitted log units in
mean level (95% simulation interval 0.093 to 1.469; simulation-tail p = 0.028)
and by 1.455 AUC units (0.180 to 2.845; p = 0.028). Its fitted peak value is
also higher by 0.884 units (0.170 to 1.588; p = 0.016). These are prespecified
moment summaries, not independent discoveries; after correcting across every
moment, window, and feeding comparison in the complete report, their q-values
are 0.311. Most importantly, the breastfed and mixed maxima occur at the end of
the 3-to-5 month window and are labeled `boundary_only`. The model supports
continued rise through the window, not an estimated biological peak at month 5.

# Retain fits for trajectory comparisons

Once the screen has ranked features, refit a small follow-up set with retained
fits. This is the bridge from "which features are significant?" to "which
trajectories look coordinated?"

```{r retained-fits, eval = FALSE}
follow_up_features <- head(
  ranked_features$feature[
    ranked_features$status == "ok" & ranked_features$p_adjust < 0.05
  ],
  8
)

feeding_4mo_followup <- test_assay_features(
  ds_4mo,
  assay = "taxa",
  features = follow_up_features,
  subject = "subject",
  time = "ageMonths",
  covariates = list(
    group_effect("feeding_4mo", role = "derived_exposure")
  ),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 6,
    na_action = "explicit",
    sparse_min_n = 5
  ),
  hypothesis = "trajectory_difference",
  test_covariate = "feeding_4mo",
  test_levels = c("ExclBreastFed", "ExclFormulaFed", "Mixed"),
  test_time_values = seq(3, 5, length.out = 30),
  test_simulations = 1999,
  test_seed = 1,
  keep_fits = TRUE,
  error_action = "continue"
)
```

The retained follow-up run uses the first eight FDR-significant taxa. The fitted
features were:

```{r observed-followup-features}
observed_followup_features <- data.frame(
  feature = c(
    "Sellimonas_intestinalis",
    "Flavonifractor_plautii",
    "Granulicatella_SGB8255",
    "Lacticaseibacillus_rhamnosus",
    "Actinomyces_SGB17168",
    "Enterococcus_casseliflavus",
    "Actinomyces_sp_HMSC065F12",
    "TM7_phylum_sp_oral_taxon_352"
  ),
  q_value = rep(0.01375, 8),
  max_separation_month = c(4.17, 4.38, 4.24, 4.66, 4.03, 4.10, 4.45, 4.03)
)

observed_followup_features
```

# Compare trajectories

`compare_trajectories()` returns the analysis layer biologists and
biostatisticians will ask for: a long data frame. Each row is a feature pair,
a metadata stratum, a time window, a metric, and its descriptive evidence.

```{r compare-trajectories, eval = FALSE}
trajectory_comparison <- compare_trajectories(
  feeding_4mo_followup,
  n = 30,
  metrics = c(
    "fitted_correlation",
    "fitted_distance",
    "derivative_correlation"
  )
)

head(trajectory_comparison)
```

In the corrected follow-up, the comparison table had 336 rows across the
exclusive-breastfed, exclusive-formula-fed, mixed, and unknown `feeding_4mo`
groups. The smallest fitted distances in the exclusive-breastfed stratum link
`Sellimonas_intestinalis` with `Flavonifractor_plautii`; the same pair is also
nearly coincident in the mixed-feeding stratum. `Sellimonas` and the TM7 feature
form another close exclusive-breastfeeding pair. These are concrete
co-trajectory hypotheses for follow-up, not proof that the organisms interact
or share a mechanism.

The p-values in this table are deliberately labeled as descriptive. The fitted
grid points are not independent biological observations.

# Generate a distance matrix

Heatmap inputs are derived from the long comparison table. This keeps the
statistical evidence table as the source of truth and lets plotting code choose
which group and time window to visualize.

```{r distance-matrix, eval = FALSE}
feeding_4mo_distance <- trajectory_distance_matrix(
  trajectory_comparison
)

feeding_4mo_distance
```

The completed run produced an 8-by-8 fitted-distance matrix for the first
available fitted-distance stratum, the exclusive-breastfed trajectory group.
The matrix is heatmap-ready, but its biological reading still depends on the
chosen stratum and the stability of that stratum. The long table remains the
source of truth when a specific feeding group or time window is required.

# Interpretation

This vignette supports three conclusions.

First, deriving `feeding_4mo` changes the question in a useful way. The model is
no longer asking only whether the sample's current feeding state matches the
sample's current abundance. It asks whether early feeding history stratifies the
trajectory.

Second, the result is selective rather than global. Twelve features support
feeding-window trajectory separation at q < 0.05, and `B. breve` reaches the
q < 0.10 tier. `B. longum` supports the targeted exclusive-breastfeeding versus
formula contrast but not the broader three-group assay-wide discovery claim.

Third, the corrected 3-to-5 month window materially improves exposure support.
Unknown `feeding_4mo` is no longer the dominant stratum, although exclusive
formula feeding remains smaller than exclusive breastfeeding. Those levels
should remain in the model frame so samples are not silently discarded, but
formula-specific claims should still be treated as follow-up hypotheses.
