---
title: "Gaussian-process follow-up for screened trajectories"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gaussian-process follow-up for screened trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

# Overview

The Gaussian-process backend is a follow-up option for features nominated by
the all-feature screen. It uses the same parse, specification, model-frame, fit,
and summary workflow as spline models. The difference is the model engine:
`trajectory_control(engine = "gp")`.

Fitted GP models require `brms` and a working Stan backend. The fitting chunks
below are not evaluated during vignette builds.

# Prepare example data

```{r load-example}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

shared_samples <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]

metadata$subject <- factor(metadata$subject)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)
```

# Declare a GP model frame for a screened feature

```{r gp-frame, eval = FALSE}
ranked_features <- feature_tests$table[
  order(feature_tests$table$p_adjust, feature_tests$table$p_value),
]
feature <- ranked_features$feature[ranked_features$status == "ok"][[1]]

spec <- model_spec(
  response = assay_feature("taxa", feature),
  subject = subject_var("subject"),
  time = time_var("ageMonths")
)

gp_mf <- model_frame(
  ds,
  trajectory(spec),
  control = trajectory_control(
    engine = "gp",
    gp_kernel = "matern32",
    transform = "log1p",
    na_action = "explicit",
    sparse_min_n = 5,
    chains = 4,
    iter = 2000,
    cores = 4,
    seed = 1
  )
)

gp_mf
```

# Fit and summarize

```{r gp-fit, eval = FALSE}
gp_fit <- fit_trajectory(gp_mf)

trajectory_results(gp_fit)
trajectory_effects(gp_fit)
```

# Kernel options

The starting GP kernels are:

```{r gp-kernels, eval = FALSE}
trajectory_control(engine = "gp", gp_kernel = "rbf")
trajectory_control(engine = "gp", gp_kernel = "matern32")
```
