---
title: "Fitting and interpreting a real-data Gaussian process"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting and interpreting a real-data Gaussian process}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
run_gp <- identical(Sys.getenv("LEVAIM_RUN_GP_VIGNETTE"), "true")
```

```{r setup}
library(LEVAiM)
```

# Follow a biological benchmark through the GP backend

The assay-wide spline screen identifies candidates efficiently. A Gaussian
process is a more expensive follow-up model for a trajectory whose uncertainty
and local rates of change deserve posterior treatment. Here the benchmark is
*Bifidobacterium longum*, an early gut colonizer already examined in the feeding
workflow.

This vignette executes the complete GP path:

`parse -> specify -> build -> compile -> fit -> diagnose -> predict -> derive -> summarize`

The analysis uses every aligned Backhed sample with an observed response. GP
fitting is opt-in during routine package builds because recompiling a full
posterior is expensive. Set `LEVAIM_RUN_GP_VIGNETTE=true` to execute the fitted
analysis. A separately labeled automated smoke test uses tiny data to verify
backend mechanics; it is not biological evidence.

# Parse the Backhed profile

```{r load-example}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

shared_samples <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]
metadata$subject <- factor(metadata$subject)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)

b_longum <- grep("s__Bifidobacterium_longum$", colnames(taxa), value = TRUE)
stopifnot(length(b_longum) == 1L)
```

# Specify, build, and compile

The model uses a Matern-3/2 covariance and a subject-specific intercept. The
response is `log1p` relative abundance, matching the spline follow-up scale.

```{r gp-frame}
spec <- model_spec(
  response = assay_feature("taxa", b_longum),
  subject = subject_var("subject"),
  time = time_var("ageMonths")
)

gp_mf <- model_frame(
  ds,
  trajectory(spec),
  control = trajectory_control(
    engine = "gp",
    gp_kernel = "matern32",
    transform = "log1p",
    chains = 4,
    iter = 2000,
    cores = 4,
    adapt_delta = 0.999,
    max_treedepth = 15,
    seed = 11
  )
)

gp_mf$formula
```

# Fit and inspect diagnostics

```{r gp-fit, eval = run_gp}
gp_fit <- suppressWarnings(fit_trajectory(gp_mf))
gp_results <- suppressWarnings(trajectory_results(gp_fit))

gp_results[c("n", "r_sq", "deviance_explained")]
gp_results$smooth_terms
gp_results$diagnostics
stopifnot(isTRUE(gp_results$diagnostics$convergence_ok))
```

The workflow does not proceed to biological interpretation unless this table
reports acceptable R-hat, effective sample sizes, and zero divergences.

# Predict and display the trajectory

```{r gp-effects, fig.height = 4.2, fig.width = 7, eval = run_gp}
gp_effects <- trajectory_effects(gp_fit, n = 50)
head(gp_effects)

plot_trajectory_report(gp_fit, n = 50)
```

# Derive posterior rates of change

`trajectory_derivatives()` differentiates each posterior fitted trajectory on
the requested time grid. Its interval is therefore an interval for the local
rate itself, not the derivative of an upper or lower fitted-curve band.

```{r gp-derivatives, eval = run_gp}
gp_derivatives <- trajectory_derivatives(
  gp_fit,
  n = 40,
  draws = 200,
  level = 0.95,
  seed = 12
)

gp_derivatives[
  , c(
    "ageMonths", ".derivative", ".lower_derivative",
    ".upper_derivative", ".probability_positive", ".sign"
  )
]
```

`increasing` means the entire pointwise credible interval is above zero;
`decreasing` means it is below zero; `uncertain` means zero remains compatible
with the posterior derivative. These are local model summaries, not automatic
colonization phases or tests of a biological mechanism.

In this full-cohort fit, the posterior rate is most likely positive during the
early months and turns predominantly negative later, but every 95% pointwise
derivative interval includes zero. The model therefore supports a broad rise and
fall in fitted abundance without claiming a sharply localized growth or decline
phase.

# Summarize biologically declared windows

```{r gp-moments, fig.height = 4.5, fig.width = 7, eval = run_gp}
gp_moments <- trajectory_moments(
  gp_fit,
  n = 80,
  windows = data.frame(
    label = c("early", "feeding_window", "later"),
    start = c(0, 3, 5),
    end = c(3, 5, 12)
  ),
  draws = 200,
  seed = 13
)

gp_moments[
  gp_moments$moment %in% c("mean_level", "auc", "linear_trend"),
]

plot_trajectory_moments(
  gp_moments,
  moment = c("mean_level", "auc", "linear_trend")
)
```

# Serialize the fitted workflow

```{r gp-serialization, eval = run_gp}
gp_path <- tempfile(fileext = ".rds")
save_levaim_object(gp_fit, gp_path)
restored_gp <- load_levaim_object(gp_path)

head(predict(restored_gp, newdata = gp_mf$data))
```

# Refit for scientific inference

If the initial full-cohort fit fails its diagnostic gate, increase MCMC depth or
revise the specification before reporting posterior derivatives or moments.

```{r production-controls, eval = FALSE}
production_control <- trajectory_control(
  engine = "gp",
  gp_kernel = "matern32",
  transform = "log1p",
  chains = 4,
  iter = 4000,
  cores = 4,
  adapt_delta = 0.999,
  max_treedepth = 15,
  seed = 11
)
```

RBF remains available through `gp_kernel = "rbf"`. Kernel choice is a model
assumption and should be checked with diagnostics and subject-grouped predictive
validation rather than selected from the smoothest-looking curve.

# Approximate GPs, additional kernels, and response families

Set `gp_basis_k` to use the Hilbert-space approximation implemented by
`brms::gp(k = ...)`. Leaving it `NULL` retains the exact GP. The basis dimension
is an approximation control and should be checked for sensitivity; it is not a
generic sparsity threshold for the microbiome assay.

```{r gp-expanded-controls, eval = FALSE}
approximate_control <- trajectory_control(
  engine = "gp",
  gp_kernel = "matern52",
  gp_basis_k = 20,
  chains = 4,
  iter = 2000
)

rougher_control <- trajectory_control(
  engine = "gp",
  gp_kernel = "exponential"
)
```

The complete covariance set is RBF, Matern-1/2 (exponential), Matern-3/2, and
Matern-5/2. Gaussian, binomial, and beta response families share the same GP
trajectory API. Binomial responses must be 0/1, and beta responses must remain
strictly inside `(0, 1)`.

```{r gp-families, eval = FALSE}
occurrence_control <- trajectory_control(
  engine = "gp",
  family = "binomial",
  gp_basis_k = 20
)

positive_abundance_control <- trajectory_control(
  engine = "gp",
  family = "beta",
  gp_basis_k = 20
)
```

Multiple covariates are declared in the model specification exactly as they are
for splines; the response remains one assay feature at a time.

```{r gp-multiple-covariates, eval = FALSE}
adjusted_spec <- model_spec(
  response = assay_feature("taxa", b_longum),
  subject = subject_var("subject"),
  time = time_var("ageMonths"),
  covariates = list(
    group_effect("feeding_4mo", role = "derived_exposure"),
    continuous_effect("maternal_age")
  )
)
```
