---
title: "A crash course in Gaussian-process trajectories"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A crash course in Gaussian-process trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

# Begin with the problem, not the acronym

Imagine a scatterplot of microbial abundance against infant age. The dots do
not arrive with a curve attached. We want to describe the underlying trajectory,
but we do not know whether it is straight, rises and plateaus, peaks, falls, or
changes direction more than once.

Ordinary linear regression makes a strong opening decision: the trajectory is a
line. Polynomial regression chooses a larger but still rigid family of shapes.
Spline regression joins flexible basis functions. A Gaussian process takes a
different route: it starts with a distribution over possible functions.

In plain language, the model says:

> Before seeing the data, many curves are possible. Nearby ages should usually
> have related values. After seeing the observations, curves that agree with the
> data become more plausible and curves that disagree become less plausible.

The result is not just one fitted line. It is a posterior distribution of
curves. The central fitted trajectory summarizes that distribution, while its
spread records what remains uncertain.

# So what is a Gaussian process?

The name contains two useful clues.

**Process** means that the object of interest is a function evaluated across an
input domain. Here the input is time and the function is a latent microbial
trajectory. We might ask for its value at birth, four months, 4.1 months, or any
other age inside the modeled range.

**Gaussian** does not necessarily mean that the observed microbiome measurements
must follow a Gaussian distribution. It means that, before conditioning on the
data, any finite collection of latent function values is modeled jointly with a
multivariate normal distribution. For example,

\[
\{f(1), f(4), f(12)\}
\]

has a joint Gaussian distribution whose covariance says how strongly the
trajectory values at 1, 4, and 12 months should be related.

This is the key move. Instead of putting a normal distribution on one number, a
GP uses a coherent family of multivariate normal distributions to put a
distribution over an entire function.

# How does one age inform another?

Suppose we observed a high abundance at four months. That observation should
usually tell us quite a lot about 4.1 months, somewhat less about five months,
and much less about twelve months. The **kernel** formalizes that decay in
relatedness.

The GP combines three ingredients:

1. a mean function describing the starting expectation;
2. a kernel describing how function values co-vary across time;
3. observations that update the prior distribution of curves into a posterior
   distribution of curves.

Where observations are dense and consistent, the posterior trajectory can be
precise. Between distant or sparse observations, more curves remain plausible
and uncertainty widens. The GP is flexible because it learns from temporal
proximity; it is not unconstrained.

# The model LEVAiM fits

For subject \(i\) at time \(t\), a LEVAiM GP can be read as

\[
g\{E(y_{it})\} = X_{it}\beta + f(t) + b_i,
\qquad
f \sim \operatorname{GP}(0, K).
\]

Read the pieces from left to right:

- \(y_{it}\) is the observed response for subject \(i\) at time \(t\);
- \(g\) is the link function associated with the response family;
- \(X_{it}\beta\) contains declared feeding, delivery, clinical, or technical
  covariate effects;
- \(f(t)\) is the shared time trajectory governed by the GP;
- \(b_i\) allows each subject to have a different baseline;
- \(K\) is the kernel that controls how strongly one time informs another.

The notation \(f \sim \operatorname{GP}(0,K)\) means that the prior mean of the
latent function is zero after accounting for the other model terms, while its
covariance across time is determined by \(K\). It does not mean the fitted
trajectory must remain near zero.

# What the GP does, and does not, decide

The GP contributes a structured prior over the unknown function \(f(t)\). It
does not erase the rest of the model. Three decisions remain distinct:

1. **Likelihood:** what kind of response was observed?
2. **Mean structure:** which feeding, delivery, clinical, or technical
   covariates belong in the model?
3. **Covariance kernel:** how should information be shared across time?

Confusing these layers produces confident nonsense. A smoother kernel cannot
repair the wrong response family, and Bayesian fitting cannot repair an exposure
that was derived from the wrong biological window.

# Kernels are assumptions about temporal relatedness

For two times \(t\) and \(t'\), a stationary kernel depends on their distance
\(r = |t-t'|\). LEVAiM exposes four `brms` covariance choices:

| LEVAiM name | Covariance | Practical character |
| --- | --- | --- |
| `"rbf"` | Exponentiated quadratic | Very smooth trajectories |
| `"matern52"` | Matern 5/2 | Smooth, but less rigid than RBF |
| `"matern32"` | Matern 3/2 | Moderately rough; the starting default |
| `"exponential"` | Matern 1/2 | Roughest supported trajectories |

The following plot shows correlation decay at a common length scale. It is a
picture of the model assumption, not a fitted microbiome result.

```{r kernel-shapes, fig.height = 4.2, fig.width = 7}
distance <- seq(0, 1, length.out = 200)
length_scale <- 0.2
scaled <- distance / length_scale

kernel_shapes <- data.frame(
  distance = rep(distance, 4),
  correlation = c(
    exp(-0.5 * scaled^2),
    (1 + sqrt(5) * scaled + 5 * scaled^2 / 3) * exp(-sqrt(5) * scaled),
    (1 + sqrt(3) * scaled) * exp(-sqrt(3) * scaled),
    exp(-scaled)
  ),
  kernel = factor(
    rep(c("RBF", "Matern 5/2", "Matern 3/2", "Exponential"), each = length(distance)),
    levels = c("RBF", "Matern 5/2", "Matern 3/2", "Exponential")
  )
)

ggplot2::ggplot(
  kernel_shapes,
  ggplot2::aes(x = distance, y = correlation, color = kernel)
) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::scale_color_manual(
    values = c("#1B6CA8", "#378B29", "#B05A00", "#8A3FFC")
  ) +
  ggplot2::labs(
    x = "Scaled temporal distance",
    y = "Prior correlation",
    color = "Kernel"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

The **length scale** determines how quickly this correlation decays. A large
length scale favors slow temporal change; a small one permits rapid change. The
GP marginal standard deviation controls how far the function can move from its
mean. Both are learned from the data, conditional on the kernel and priors.

# Irregular observations are not missing-data magic

GPs naturally predict on a continuous time axis even when subjects were sampled
at different ages. A child does not need an observation at every prediction
month. Information is borrowed through the covariance kernel, and uncertainty
usually increases where temporal support is weak.

That does **not** mean a GP automatically resolves every missing value:

- an unobserved response at a time point need not be fabricated before fitting;
- a missing covariate such as feeding state still lacks a model input;
- an unknown exposure history still requires explicit derivation, imputation,
  or an uncertainty model;
- a structurally absent taxon is not interchangeable with a missing assay
  measurement.

LEVAiM therefore handles irregular chronology through the GP while keeping
missing-annotation policy explicit in preprocessing and the model specification.

# The likelihood defines the question

The same temporal GP can sit inside different response models:

- `family = "gaussian"` models a continuous response, optionally transformed;
- `family = "binomial"` models the probability of a 0/1 event such as taxon
  detection;
- `family = "beta"` models a continuous proportion strictly inside `(0, 1)`.

Zeros are not silently nudged into a beta model. For zero-containing relative
abundance, occurrence and positive abundance answer different questions and may
be modeled separately. See
`vignette("response-families", package = "LEVAiM")`.

# Exact and approximate GPs

An exact GP works with a covariance matrix whose computational cost grows
quickly with the number of observations. LEVAiM uses the exact path when
`gp_basis_k = NULL`.

```{r exact-control}
exact_gp <- trajectory_control(
  engine = "gp",
  gp_kernel = "matern32",
  gp_basis_k = NULL
)
```

Setting `gp_basis_k` requests the Hilbert-space approximation exposed by
`brms::gp(k = ...)`. The latent function is represented with a finite basis,
reducing the cost of larger fits.

```{r approximate-control}
approximate_gp <- trajectory_control(
  engine = "gp",
  gp_kernel = "matern32",
  gp_basis_k = 20
)
```

The basis dimension is not a universal tuning constant. Too few functions can
erase real temporal structure; more functions cost more. A scientific analysis
should check whether conclusions are stable across defensible basis dimensions
and, when feasible, against an exact fit.

# Posterior fitting is only useful when it behaves

`brms` uses Stan to sample the posterior. A returned model object is not proof
that the posterior was explored adequately. LEVAiM reports:

- maximum R-hat;
- minimum bulk and tail effective sample sizes;
- divergent transitions;
- a combined convergence gate.

The POC gate requires R-hat no greater than 1.01, bulk and tail ESS of at least
400, and zero divergences. Failure means refit or revise the model before
interpreting trajectories. Increasing `adapt_delta` can address occasional
divergences, but persistent failures call for prior or model reconsideration.

# Derivatives answer local questions

A fitted value asks, “what level does the trajectory have here?” A derivative
asks, “how quickly is it changing here?” For GP fits,
`trajectory_derivatives()` differentiates every posterior fitted curve and then
summarizes those rates.

```{r derivative-api, eval = FALSE}
rates <- trajectory_derivatives(
  gp_fit,
  time_values = seq(0, 12, by = 0.25),
  draws = 500,
  level = 0.95
)
```

The output includes the posterior mean derivative, credible interval,
`P(derivative > 0)`, and a pointwise sign label. A fitted rise followed by a fall
does not guarantee that every local derivative excludes zero. Nor does every
trajectory need to contain an interior peak or valley.

# From one curve to biological evidence

The shared LEVAiM sequence is deliberately explicit:

```{r workflow-api, eval = FALSE}
spec <- model_spec(
  response = assay_feature("taxa", feature),
  subject = subject_var("subject"),
  time = time_var("ageMonths"),
  covariates = list(group_effect("feeding_4mo", role = "derived_exposure"))
)

mf <- model_frame(
  ds,
  trajectory(spec, varying_trajectory("feeding_4mo")),
  control = trajectory_control(
    engine = "gp",
    gp_kernel = "matern32",
    chains = 4,
    iter = 2000,
    adapt_delta = 0.99
  )
)

fit <- fit_trajectory(mf)
results <- trajectory_results(fit)
effects <- trajectory_effects(fit)
rates <- trajectory_derivatives(fit)
moments <- trajectory_moments(fit)
```

The GP describes trajectories conditional on that specification. Biological
evidence still depends on the contrast being asked, the exposure definition,
temporal support, multiplicity, diagnostics, and whether the observational
design can distinguish association from mechanism.

# Words you will see in the output

| Term | Meaning in this workflow |
| --- | --- |
| Observation | One measured response from one sample at one recorded time |
| Latent trajectory | The unobserved smooth function the model is trying to learn |
| Prior | Which trajectories and parameter values were plausible before using these observations |
| Posterior | Which trajectories remain plausible after combining the prior with the observations |
| Posterior mean | The average fitted trajectory across posterior draws |
| Credible interval | A posterior uncertainty interval for a fitted value, derivative, or summary |
| Kernel | The rule describing how strongly function values at different times co-vary |
| Length scale | The temporal distance over which the trajectory remains strongly related |
| GP marginal SD | How much the latent trajectory is allowed to vary around its mean |
| Posterior draw | One jointly plausible realization of the model parameters and trajectory |

A 95% credible interval is a statement about the fitted posterior under the
specified model. It is not a guarantee that 95% of future observations will lie
inside that interval, and it does not turn an observational association into a
causal effect.

# A compact reading checklist

Before believing a GP trajectory, ask:

1. Does the response family match the measurement and its boundaries?
2. Was the exposure derived from a biologically defensible window?
3. Which kernel was assumed, and are conclusions robust to reasonable choices?
4. Does the observed chronology support the region being interpreted?
5. Did R-hat, ESS, and divergence checks pass?
6. Are derivative and moment statements supported by posterior intervals?
7. Is the claim predictive, associational, or mechanistic?

Continue with `vignette("gaussian-processes", package = "LEVAiM")` for the
executed full-cohort Backhed workflow.
