---
title: "02. Turn the biological question into a model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02. Turn the biological question into a model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

**Course position:** Lecture 2 of 9<br>
**Previously:** `vignette("data-loading", package = "LEVAiM")`<br>
**Next:** `vignette("feeding-4mo-trajectories", package = "LEVAiM")`

# Today we decide what the model is allowed to mean

The biological question is deliberately specific:

> Which taxonomic features show trajectories that differ according to feeding
> history around four months, defined from observations between months 3 and 5?

LEVAiM separates that question into two explicit layers. First, derive
history-aware metadata from a time-varying exposure when the question is about
feeding history. Second, declare model covariates with roles so `model_frame()`
can validate how those columns behave within subjects.

# Prepare the BackhedF 2015 example

```{r load-example}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}
if (!dir.exists(input_dir)) {
  input_dir <- system.file("extdata", package = "LEVAiM")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

shared_samples <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]

metadata$subject <- factor(metadata$subject)
metadata$feeding_state <- factor(metadata$feeding_state)
metadata$anno_uncertain <- factor(metadata$anno_uncertain)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)
```

# Derive feeding-history features

`feeding_state` is time-varying. Modeling it directly asks about the current
observed state at each sample. If the question is about earlier feeding
history, derive subject-level features first.

```{r derive-feeding}
ds_history <- derive_time_varying_features(
  ds,
  exposure = "feeding_state",
  summaries = list(
    feeding_4mo = exposure_window(3, 5, method = "modal")
  )
)

exposure_derivations(ds_history)
```

# Define the assay-wide universe

The response rotates across every taxon in the prespecified eligible pool. We
do not preselect bifidobacteria, *Bacteroides*, or any other expected colonizer;
known biology enters after the screen, when we interpret the table.

```{r response-universe}
assay <- "taxa"
prevalence <- colMeans(taxa > 0, na.rm = TRUE)
eligible_features <- names(prevalence[prevalence >= 0.20])

length(eligible_features)
head(eligible_features)
```

# Encode the reusable model pieces

The per-feature wrapper builds one `model_spec()` per feature internally. The
stable pieces are subject, time, exposure, trajectory design, and controls.

```{r specification}
subject <- subject_var("subject")
time <- time_var("ageMonths")

history_feeding_covariates <- list(
  group_effect("feeding_4mo", role = "derived_exposure")
)

history_feeding_design <- varying_trajectory("feeding_4mo")

control <- trajectory_control(
  engine = "spline",
  transform = "log1p",
  spline_k = 3,
  na_action = "explicit",
  missing_level = "Unknown",
  sparse_min_n = 5
)
```

The cohort has three biological sampling waves. We therefore cap the spline
basis at three rather than treating small differences in recorded age as new
longitudinal support.

# Declare the trajectory contrast

The key move is `varying_trajectory("feeding_4mo")`: we are asking for
feeding-history-specific time curves, not just a single average curve with a
feeding covariate tacked on.

```{r trajectory-design}
prototype_spec <- model_spec(
  response = assay_feature(assay, eligible_features[[1]]),
  subject = subject,
  time = time,
  covariates = history_feeding_covariates
)

shared_traj <- trajectory(prototype_spec, design = naive_trajectory())
feeding_traj <- trajectory(prototype_spec, design = history_feeding_design)

shared_traj
feeding_traj
```

# Build the model frame

Missing feeding annotations are kept as an explicit level by default. Sparse
levels are collapsed before model fitting. That choice is intentional: uncertain
metadata are part of real infant cohorts, not a reason to throw away otherwise
informative longitudinal samples.

```{r model-frame}
prototype_mf <- model_frame(
  ds_history,
  feeding_traj,
  control = control
)

prototype_mf
prototype_mf$formula
prototype_mf$preprocessing$covariate_roles
```

# Pass the baton

Lecture 3 applies this exact template across all eligible species. Nothing
about organism identity enters the model factory or multiplicity family.

```{r next-lecture, eval = FALSE}
vignette("feeding-4mo-trajectories", package = "LEVAiM")
```
