---
title: "Specifying an assay-wide feeding trajectory screen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specifying an assay-wide feeding trajectory screen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

# Objective

The biological question is deliberately specific:

> Which taxonomic features show trajectories that differ by feeding state over
> infancy?

LEVAiM separates that question into two explicit layers. First, derive
history-aware metadata from a time-varying exposure when the question is about
feeding history. Second, declare model covariates with roles so `model_frame()`
can validate how those columns behave within subjects.

# Prepare the BackhedF 2015 example

```{r load-example}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

shared_samples <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]

metadata$subject <- factor(metadata$subject)
metadata$feeding_state <- factor(metadata$feeding_state)
metadata$anno_uncertain <- factor(metadata$anno_uncertain)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)
```

# Derive feeding-history features

`feeding_state` is time-varying. Modeling it directly asks about the current
observed state at each sample. If the question is about earlier feeding
history, derive subject-level features first.

```{r derive-feeding}
ds_history <- derive_time_varying_features(
  ds,
  exposure = "feeding_state",
  summaries = list(
    feeding_4mo = exposure_window(3, 5, method = "modal"),
    feeding_6mo = exposure_window(5, 7, method = "modal"),
    ever_exclusive_breastfed = exposure_ever("ExclBreastFed"),
    feeding_last_observed = exposure_last_observed(),
    feeding_transition = exposure_transition()
  )
)

exposure_derivations(ds_history)
```

# Define the assay-wide hypothesis

The response will rotate across every taxonomic feature in the assay. We do not
preselect bifidobacteria, Bacteroides, or any other expected colonizer. The
known biology enters after the screen, when we interpret the result table.

```{r response-universe}
assay <- "taxa"
all_features <- colnames(taxa)

length(all_features)
head(all_features)
```

# Encode the reusable model pieces

The per-feature wrapper builds one `model_spec()` per feature internally. The
stable pieces are subject, time, covariates, trajectory design, and controls. A
current-state screen and a history-derived screen are different biological
questions, so they use different covariates.

```{r specification}
subject <- subject_var("subject")
time <- time_var("ageMonths")

current_feeding_covariates <- list(
  group_effect("feeding_state", role = "time_varying"),
  group_effect("anno_uncertain")
)

history_feeding_covariates <- list(
  group_effect("feeding_4mo", role = "derived_exposure"),
  group_effect("ever_exclusive_breastfed", role = "derived_exposure"),
  group_effect("feeding_transition", role = "derived_exposure")
)

current_feeding_design <- varying_trajectory("feeding_state")
history_feeding_design <- varying_trajectory("feeding_4mo")

control <- trajectory_control(
  engine = "spline",
  transform = "log1p",
  na_action = "explicit",
  missing_level = "Unknown",
  sparse_min_n = 5
)
```

# Declare the trajectory contrast

The key move is `varying_trajectory("feeding_state")`: we are asking for
feeding-state-specific time curves, not just a single average curve with a
feeding covariate tacked on.

```{r trajectory-design}
prototype_spec <- model_spec(
  response = assay_feature(assay, all_features[[1]]),
  subject = subject,
  time = time,
  covariates = history_feeding_covariates
)

shared_traj <- trajectory(prototype_spec, design = naive_trajectory())
feeding_traj <- trajectory(prototype_spec, design = history_feeding_design)

shared_traj
feeding_traj
```

# Build the model frame

Missing feeding annotations are kept as an explicit level by default. Sparse
levels are collapsed before model fitting. That choice is intentional: uncertain
metadata are part of real infant cohorts, not a reason to throw away otherwise
informative longitudinal samples.

```{r model-frame}
prototype_mf <- model_frame(
  ds_history,
  feeding_traj,
  control = control
)

prototype_mf
prototype_mf$preprocessing$covariate_roles
```

# Continue

Use `vignette("feature-testing", package = "LEVAiM")` to apply this model
template across every taxonomic feature.
