---
title: "07. Match the model to the response"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{07. Match the model to the response}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

**Course position:** Lecture 7 of 9<br>
**Previously:** `vignette("trajectory-analysis", package = "LEVAiM")`<br>
**Next:** `vignette("gp-crash-course", package = "LEVAiM")`

# Ask the response a question it can answer

Microbial relative abundance is bounded, often zero, and rarely Gaussian on its
original scale. A response family is therefore part of the biological
hypothesis, not a cosmetic fitting option.

This vignette uses *Bifidobacterium longum* to separate two questions:

1. **Occurrence:** does the probability of detecting the organism change with
   age? This is a binomial response containing exactly 0 and 1.
2. **Positive abundance:** among samples where the organism is detected, how
   does its relative abundance change with age? This is a beta response strictly
   between 0 and 1.

These models do not jointly constitute a zero-inflated beta model. They are an
explicit two-part analysis with separately interpretable estimands.
Both are age-only models so we can isolate the response-family choice. A
feeding analysis would add the derived `feeding_4mo` exposure and trajectory
design from Lectures 2–4 to each part separately.

# Load the longitudinal profile

```{r load-data}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}
if (!dir.exists(input_dir)) {
  input_dir <- system.file("extdata", package = "LEVAiM")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

shared_samples <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]
metadata$subject <- factor(metadata$subject)

b_longum <- grep("s__Bifidobacterium_longum$", colnames(taxa), value = TRUE)
stopifnot(length(b_longum) == 1L)

# MetaPhlAn relative abundance is expressed as a percentage in this file.
abundance <- taxa[, b_longum]
metadata$b_longum_detected <- as.integer(abundance > 0)
metadata$b_longum_positive_fraction <- ifelse(
  abundance > 0,
  abundance / 100,
  NA_real_
)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)
```

# Model detection probability

The binomial model retains both detections and non-detections. Predictions are
probabilities on the response scale, and subject-grouped validation defaults to
log loss and Brier score.

```{r binomial-fit}
occurrence_spec <- model_spec(
  response = metadata_feature("b_longum_detected"),
  subject = subject_var("subject"),
  time = time_var("ageMonths")
)

occurrence_mf <- model_frame(
  ds,
  trajectory(occurrence_spec),
  control = trajectory_control(
    engine = "spline",
    family = "binomial",
    transform = "identity",
    spline_k = 6
  )
)

occurrence_fit <- fit_trajectory(occurrence_mf)
head(trajectory_effects(occurrence_fit, n = 20))
```

```{r binomial-validation}
occurrence_cv <- cross_validate_trajectory(
  occurrence_mf,
  v = 3,
  seed = 1
)
cv_summary(occurrence_cv)
```

The fitted detection probability rises from about 0.74 at birth to 0.89 by
roughly 3.5 months. Three-fold subject-grouped validation gives log loss near
0.42 and Brier score near 0.13. Those values are usable diagnostics, but without
a prevalence-only comparator they are not evidence that the smooth improves
out-of-subject prediction.

# Model abundance conditional on detection

The beta model excludes non-detections because their conditional positive
abundance is undefined. It models a fraction in the open interval `(0, 1)` and
uses the beta likelihood directly; LEVAiM does not add pseudocounts or squeeze
boundary values inward.

```{r beta-fit}
positive_spec <- model_spec(
  response = metadata_feature("b_longum_positive_fraction"),
  subject = subject_var("subject"),
  time = time_var("ageMonths")
)

positive_mf <- model_frame(
  ds,
  trajectory(positive_spec),
  control = trajectory_control(
    engine = "spline",
    family = "beta",
    transform = "identity",
    spline_k = 6
  )
)

positive_fit <- fit_trajectory(positive_mf)
head(trajectory_effects(positive_fit, n = 20))
```

```{r beta-validation}
positive_cv <- cross_validate_trajectory(
  positive_mf,
  v = 3,
  seed = 1
)
cv_summary(positive_cv)
```

Among detected samples, fitted positive abundance rises from about 0.09 at
birth to 0.18 by roughly 3.5 months. Grouped-CV RMSE is about 0.18 on the
fraction scale. This is appreciable error for a bounded response and should not
be hidden behind the cleaner-looking fitted mean.

# Interpret the two trajectories together

The occurrence trajectory estimates when *B. longum* is likely to be present.
The positive-abundance trajectory estimates its expected proportional abundance
after presence has already been established. A feeding effect could enter one,
the other, or both; those outcomes imply different colonization mechanisms.

Do not replace zero abundances with an arbitrary epsilon merely to make the beta
model run. That changes the response and conflates absence with low positive
abundance. When a single likelihood must jointly model boundary mass and the
continuous interior, use an explicit hurdle or zero/one-inflated model outside
the current LEVAiM API and report that model choice.

# Use the same families with Gaussian processes

The response definitions and boundary policy do not change when the trajectory
engine changes. GP fitting maps `family = "binomial"` to a Bernoulli likelihood
and `family = "beta"` to beta regression. Exact and Hilbert-space approximate
GPs are both available.

```{r gp-family-controls}
occurrence_gp_mf <- model_frame(
  ds,
  trajectory(occurrence_spec),
  control = trajectory_control(
    engine = "gp",
    family = "binomial",
    gp_kernel = "matern32",
    gp_basis_k = 20
  )
)

positive_gp_mf <- model_frame(
  ds,
  trajectory(positive_spec),
  control = trajectory_control(
    engine = "gp",
    family = "beta",
    gp_kernel = "matern32",
    gp_basis_k = 20
  )
)

occurrence_gp_mf$formula
positive_gp_mf$formula
```

# Pass the baton

Splines and GPs can answer the same occurrence and positive-abundance
questions, but they encode temporal smoothness differently. Lecture 8 pauses
the fitting workflow and builds the GP intuition needed to make that choice.

```{r next-lecture, eval = FALSE}
vignette("gp-crash-course", package = "LEVAiM")
```
