---
title: "04. Read one spline trajectory deeply"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{04. Read one spline trajectory deeply}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(LEVAiM)
```

**Course position:** Lecture 4 of 9<br>
**Previously:** `vignette("feeding-4mo-trajectories", package = "LEVAiM")`<br>
**Next:** `vignette("validation", package = "LEVAiM")`

# A ranked row is not yet a biological story

Lecture 3 gave us one row per species. That table is excellent for discovery,
but a p-value cannot show whether groups separate gradually, cross each other,
or differ only where observations are scarce.

We now follow *Bifidobacterium longum*. It is useful here for two reasons: it is
an early-colonization benchmark, and it lets us ask whether a familiar feeding
association is visible in the actual fitted curves. It did not bypass the
assay-wide screen.

# Reconstruct the data and exposure

```{r course-data}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) input_dir <- file.path("examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- system.file("extdata", package = "LEVAiM")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)
shared <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]
metadata$subject <- factor(metadata$subject)

ds <- LongitudinalDataset(
  list(taxa = taxa), metadata, "ageMonths", "subject"
)
ds <- derive_time_varying_features(
  ds,
  exposure = "feeding_state",
  summaries = list(feeding_4mo = exposure_window(3, 5, method = "modal"))
)

b_longum <- grep("s__Bifidobacterium_longum$", colnames(taxa), value = TRUE)
stopifnot(length(b_longum) == 1L)
```

# Follow the same parse-spec-build-compile-fit sequence

```{r fit-spline}
spec <- model_spec(
  response = assay_feature("taxa", b_longum),
  subject = subject_var("subject"),
  time = time_var("ageMonths"),
  covariates = list(
    group_effect("feeding_4mo", role = "derived_exposure")
  )
)

mf <- model_frame(
  ds,
  trajectory(spec, varying_trajectory("feeding_4mo")),
  trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 3,
    na_action = "explicit",
    sparse_min_n = 5
  )
)

mf$formula
fit <- fit_trajectory(mf)
results <- trajectory_results(fit)
results
```

The formula is worth reading. Feeding group changes the time smooth itself;
this is not merely a constant feeding offset added to one shared curve.
The smooth-term p-values test time variation inside each feeding group. They do
not test whether feeding-group trajectories differ from one another; that is
the omnibus contrast reported in Lecture 3.

# Put fitted curves back beside observations

```{r trajectory-plot, fig.height = 4.5, fig.width = 7}
feeding_levels <- c("ExclBreastFed", "ExclFormulaFed", "Mixed")

plot_trajectory_report(
  fit,
  n = 80,
  levels = feeding_levels,
  window = c(3, 5)
)
```

Read the plot in layers:

1. the points show where the data actually exist;
2. the curves show conditional fitted means;
3. the ribbons show pointwise uncertainty;
4. the highlighted window shows where Lecture 3 tested separation.

A compelling curve outside the observed support is still extrapolation. A
group with few children should usually carry more uncertainty, not more faith.

# Ask where groups differ

The screen found the expected fitted ordering but weak three-group omnibus
evidence for *B. longum* (`p = 0.085`). Pointwise contrasts show where that
ordering is largest without pretending that each time point is a new
independent discovery experiment.

```{r localized-contrasts, fig.height = 4.5, fig.width = 7}
contrasts <- trajectory_contrasts(
  fit,
  time_values = seq(3, 5, length.out = 40),
  levels = feeding_levels
)
head(contrasts)
plot_trajectory_contrasts(contrasts)
```

# Turn the curve into rates of change

```{r derivatives}
rates <- trajectory_derivatives(fit, n = 60)
rates <- rates[as.character(rates$feeding_4mo) %in% feeding_levels, ]
head(rates[c(
  "ageMonths", "feeding_4mo", ".derivative",
  ".lower_derivative", ".upper_derivative", ".sign"
)])
```

A positive derivative means the fitted abundance is increasing locally; a
negative derivative means it is decreasing. `uncertain` means the interval for
the rate includes zero. LEVAiM does not force every curve to contain a peak,
valley, or developmental phase.

# Summarize the biologically named window

```{r moments, fig.height = 4.5, fig.width = 7}
moments <- trajectory_moments(
  fit,
  n = 100,
  windows = data.frame(label = "feeding_window", start = 3, end = 5),
  levels = feeding_levels,
  draws = 300,
  seed = 4
)

moments[moments$moment %in% c(
  "mean_level", "auc", "linear_trend", "peak_time"
), ]

plot_trajectory_moments(
  moments,
  moment = c("mean_level", "auc", "linear_trend"),
  window = "feeding_window"
)
```

Moments compress a function into a data frame without erasing the window or
uncertainty that gave the summary meaning. Peak times carry statuses such as
`interior_supported` or `boundary_only`; the largest endpoint is not
automatically a biological peak.

Inside the 3-to-5-month window, the fitted mean follows the biologically
familiar order: exclusive breastfeeding (2.16), mixed feeding (1.67), then
exclusive formula feeding (1.41) on the modeled `log1p` scale. The uncertainty
intervals overlap, and every window-specific linear-trend interval includes
zero. The model therefore supports an ordered descriptive pattern, not a claim
of three decisively separated trajectories or different growth rates.

# Pass the baton

We have interpreted one fitted model in-sample. Lecture 5 asks whether the same
specification behaves sensibly when complete children are held out.

```{r next-lecture, eval = FALSE}
vignette("validation", package = "LEVAiM")
```
