---
title: "Following up a screen-derived trajectory hit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Following up a screen-derived trajectory hit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

# Objective

Here we fit an inspectable trajectory for a feature nominated by the all-assay
screen. The rule is important: first test every feature with the wrapper, then
choose follow-up targets from the result table. The goal is not to bless a
favorite organism. The goal is to understand the fitted trajectory behind a
screen-derived signal.

# Prepare the example data

```{r load-example}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

shared_samples <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]

metadata$subject <- factor(metadata$subject)
metadata$feeding_state <- factor(metadata$feeding_state)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)
```

# Choose a follow-up feature from the screen

The full screen is shown in `vignette("feature-testing", package = "LEVAiM")`.
After running it, choose a feature from the ranked result table.

```{r choose-hit, eval = FALSE}
ranked_features <- feature_tests$table[
  order(feature_tests$table$p_adjust, feature_tests$table$p_value),
]

feature <- ranked_features$feature[ranked_features$status == "ok"][[1]]
feature
```

# Fit a feeding-stratified spline

```{r fit-spline, eval = FALSE}

spec <- model_spec(
  response = assay_feature("taxa", feature),
  subject = subject_var("subject"),
  time = time_var("ageMonths"),
  covariates = list(group_effect("feeding_state"))
)

mf <- model_frame(
  ds,
  trajectory(spec, design = varying_trajectory("feeding_state")),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_basis = "gam",
    spline_k = 6,
    na_action = "explicit",
    sparse_min_n = 5
  )
)

fit <- fit_trajectory(mf)
fit
```

# Summarize the fitted model

`trajectory_results()` extracts the fitted model evidence without forcing users
to parse a backend-specific object first.

```{r results, eval = FALSE}
results <- trajectory_results(fit)
results
```

# Inspect the fitted trajectory

The fitted effects table is the basic scientific object: age, feeding state,
fitted abundance, and uncertainty.

```{r effects, fig.height = 4, eval = FALSE}
effects <- trajectory_effects(fit, n = 20)
head(effects)

plot_trajectory_effects(fit, effects = effects)
```

# Estimate local change

Derivatives turn the fitted curve into a rate-of-change table. LEVAiM reports
the rate; it does not invent a peak, valley, or phase label unless the evidence
supports one downstream.

```{r derivatives, eval = FALSE}
derivatives <- trajectory_derivatives(fit, n = 20)
head(derivatives)
```

# Continue

Use `vignette("validation", package = "LEVAiM")` to ask whether the trajectory
model behaves sensibly under subject-aware held-out validation.
