---
title: "06. Compare microbial trajectories"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{06. Compare microbial trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(LEVAiM)
```

**Course position:** Lecture 6 of 9<br>
**Previously:** `vignette("validation", package = "LEVAiM")`<br>
**Next:** `vignette("response-families", package = "LEVAiM")`

# From one trajectory to a microbial program

Lecture 3 ranked the assay. Lectures 4 and 5 then slowed down to inspect and
validate one benchmark. We can now ask a different question:

> Within each feeding group, which microbial features follow similar temporal
> paths between 3 and 5 months?

This is a follow-up analysis, not a second discovery screen. We use a small
teaching panel containing a *Bifidobacterium* benchmark and five additional
taxa from the assay-wide teaching screen. The panel is declared openly so the vignette
fits six trajectories, not 2,011 mostly empty profiles.

# Reconstruct the course dataset

```{r course-data}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) input_dir <- file.path("examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- system.file("extdata", package = "LEVAiM")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

shared <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]
metadata$subject <- factor(metadata$subject)

ds <- LongitudinalDataset(list(taxa = taxa), metadata, "ageMonths", "subject")
ds <- derive_time_varying_features(
  ds,
  "feeding_state",
  list(feeding_4mo = exposure_window(3, 5, method = "modal"))
)
```

# Declare the follow-up panel

The readable labels below are resolved to full MetaPhlAn feature names. Every
member exceeds the 20% prevalence rule used for the screen.

```{r follow-up-panel}
panel_labels <- c(
  "Bifidobacterium longum",
  "Phocaeicola vulgatus",
  "Escherichia coli",
  "Veillonella parvula",
  "Streptococcus salivarius",
  "Bacteroides uniformis"
)
panel_species <- gsub(" ", "_", panel_labels, fixed = TRUE)
panel_features <- vapply(
  panel_species,
  function(species) {
    hit <- grep(paste0("s__", species, "$"), colnames(taxa), value = TRUE)
    stopifnot(length(hit) == 1L)
    hit
  },
  character(1)
)

data.frame(
  feature = panel_labels,
  prevalence = colMeans(taxa[, panel_features, drop = FALSE] > 0),
  row.names = NULL
)
```

# Fit the same model six times

`test_assay_features()` remains useful here because it guarantees that every
feature receives the same specification. `keep_fits = TRUE` retains the fitted
objects needed for pairwise trajectory comparison.

```{r panel-fits}
panel_fits <- test_assay_features(
  ds,
  assay = "taxa",
  features = unname(panel_features),
  subject = "subject",
  time = "ageMonths",
  covariates = list(
    group_effect("feeding_4mo", role = "derived_exposure")
  ),
  design = varying_trajectory("feeding_4mo"),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 3,
    na_action = "explicit",
    sparse_min_n = 5
  ),
  keep_fits = TRUE,
  error_action = "stop"
)

names(panel_fits$fits) <- panel_labels
panel_fits$table[c("feature", "status")]
```

# Produce the comparison data frame

Each row below compares one feature pair, one feeding group, one time window,
and one metric. Correlation asks whether curve shapes move together. Distance
asks how far apart those shapes are. Derivative correlation asks whether their
local rates of change agree.

```{r compare-trajectories}
trajectory_comparison <- compare_trajectories(
  panel_fits,
  time_values = seq(3, 5, length.out = 30),
  view = "same_group_level",
  metrics = c(
    "fitted_correlation",
    "fitted_distance",
    "derivative_correlation"
  )
)

display_comparison <- trajectory_comparison
display_comparison$feature_1 <- gsub(
  "_", " ", sub(".*s__", "", display_comparison$feature_1)
)
display_comparison$feature_2 <- gsub(
  "_", " ", sub(".*s__", "", display_comparison$feature_2)
)

head(display_comparison[c(
  "feature_1", "feature_2", "group_level", "metric",
  "estimate", "p_value", "q_value"
)])
```

# Treat the p-values as descriptive

Biostatisticians will look for p-values, so the table provides them. It also
labels exactly what they are: correlations among fitted grid values. Those grid
points are generated by models from shared subjects and are not independent
biological observations. The p-values and q-values describe curve alignment;
they do not test subject-level co-dynamics or a causal microbial interaction.

```{r inference-contract}
unique(trajectory_comparison[c(
  "test_method", "null_hypothesis", "inference_warning"
)])
```

# Build one distance matrix per feeding group

A single heatmap would collapse biologically distinct trajectories. We instead
extract one matrix for each feeding group from the same long table.

```{r distance-matrices}
feeding_groups <- c("ExclBreastFed", "ExclFormulaFed", "Mixed")
trajectory_labels <- unique(rbind(
  data.frame(
    trajectory_id = trajectory_comparison$trajectory_id_1,
    feature = trajectory_comparison$feature_1
  ),
  data.frame(
    trajectory_id = trajectory_comparison$trajectory_id_2,
    feature = trajectory_comparison$feature_2
  )
))
trajectory_labels$feature <- gsub(
  "_", " ", sub(".*s__", "", trajectory_labels$feature)
)

distance_matrices <- setNames(
  lapply(feeding_groups, function(feeding_group) {
    group_comparison <- trajectory_comparison[
      trajectory_comparison$group_level == feeding_group,
    ]
    distance <- trajectory_distance_matrix(group_comparison)
    readable_names <- trajectory_labels$feature[
      match(rownames(distance), trajectory_labels$trajectory_id)
    ]
    rownames(distance) <- readable_names
    colnames(distance) <- readable_names
    distance
  }),
  feeding_groups
)

names(distance_matrices)
distance_matrices[[1]]
```

These matrices are ready for heatmaps or clustering, but the matrix identity
must travel with the result: a pair of taxa can move together under one feeding
history and diverge under another.

In the exclusive-breastfed matrix, *B. longum* and *V. parvula* have nearly
identical standardized fitted shapes (distance 0.029). *S. salivarius* and
*E. coli* form another close pair (distance 0.029), while those two pairs are
almost maximally separated. Notice also that fitted-level and derivative
correlations need not agree: two curves can occupy similar levels while changing
at different local rates. These are model- and window-dependent shape summaries,
not evidence that the organisms interact or co-vary within individual children.

# Pass the baton

So far every response has been transformed continuous abundance. Lecture 7
asks whether the biology calls for a different response: detection probability
or abundance conditional on detection.

```{r next-lecture, eval = FALSE}
vignette("response-families", package = "LEVAiM")
```
