---
title: "Comparing trajectories after an all-feature screen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing trajectories after an all-feature screen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

# Overview

This vignette starts after the all-feature screen. It covers two outputs:

1. a long trajectory-comparison table from `compare_trajectories()`;
2. a fitted-trajectory distance matrix from `trajectory_distance_matrix()`.

The correlation p-values in the comparison table are descriptive summaries of
fitted grid curves. They are labeled as descriptive because fitted grid points
are not independent biological observations.

# Prepare retained feature fits from the screen

```{r feature-fits}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

shared_samples <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]

metadata$subject <- factor(metadata$subject)
metadata$feeding_state <- factor(metadata$feeding_state)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)

ds_history <- derive_time_varying_features(
  ds,
  exposure = "feeding_state",
  summaries = list(
    feeding_4mo = exposure_window(3, 5, method = "modal"),
    feeding_transition = exposure_transition()
  )
)
```

# Run the all-feature screen

The first pass tests every feature. Then a focused set of hits is refit with
`keep_fits = TRUE` so trajectories can be compared.

```{r all-feature-screen, eval = FALSE}
feature_tests <- test_assay_features(
  ds,
  assay = "taxa",
  features = colnames(taxa),
  subject = "subject",
  time = "ageMonths",
  covariates = list(group_effect("feeding_state", role = "time_varying")),
  design = varying_trajectory("feeding_state"),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 6,
    na_action = "explicit",
    sparse_min_n = 5
  )
)

ranked_features <- feature_tests$table[
  order(feature_tests$table$p_adjust, feature_tests$table$p_value),
]

follow_up_features <- head(ranked_features$feature[ranked_features$status == "ok"], 8)
```

# Refit selected screen hits with retained fits

The follow-up set is selected from the screen. It may include familiar early
colonizers, unexpected taxa, or both.

```{r retained-fits, eval = FALSE}
feature_tests_with_fits <- test_assay_features(
  ds,
  assay = "taxa",
  features = follow_up_features,
  subject = "subject",
  time = "ageMonths",
  covariates = list(group_effect("feeding_state", role = "time_varying")),
  design = varying_trajectory("feeding_state"),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 6,
    na_action = "explicit",
    sparse_min_n = 5
  ),
  keep_fits = TRUE
)
```

# Compare trajectories

`compare_trajectories()` returns a long data frame. Each row is a feature pair,
trajectory group, time window, and metric.

```{r compare, eval = FALSE}
trajectory_comparison <- compare_trajectories(
  feature_tests_with_fits,
  n = 25,
  metrics = c("fitted_correlation", "fitted_distance", "derivative_correlation")
)

head(trajectory_comparison)
```

# Inspect descriptive inference labels

```{r descriptive-labels, eval = FALSE}
trajectory_comparison[
  trajectory_comparison$metric == "fitted_correlation",
  c("feature_1", "feature_2", "group_level", "estimate", "p_value", "q_value",
    "test_method", "inference_warning")
]
```

# Generate a distance matrix

Distance matrices are derived from the long comparison table. They are suitable
inputs for heatmap code; plotting is intentionally left to downstream workflow
decisions.

```{r distance-matrix, eval = FALSE}
trajectory_distance_matrix(trajectory_comparison)
```
