---
title: "Validating a screen-derived trajectory model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating a screen-derived trajectory model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(LEVAiM)
```

# Overview

Once the all-feature screen nominates follow-up targets, validation asks whether
a fitted trajectory behaves sensibly when whole subjects are held out. Repeated
samples from the same subject must never be split across train and test sets.
This vignette covers:

1. grouped fold generation;
2. repeated grouped cross-validation;
3. aggregate validation summaries;
4. held-out prediction tables.

# Prepare example data and model frame

Run the all-feature screen first, then set `feature` from the ranked result
table. The chunk below is not evaluated during vignette builds because it
depends on that completed screen.

```{r model-frame, eval = FALSE}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- file.path("examples", "input_files")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)

shared_samples <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared_samples, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]

metadata$subject <- factor(metadata$subject)

ds <- LongitudinalDataset(
  assays = list(taxa = taxa),
  metadata = metadata,
  time_col = "ageMonths",
  subject_col = "subject"
)

ranked_features <- feature_tests$table[
  order(feature_tests$table$p_adjust, feature_tests$table$p_value),
]
feature <- ranked_features$feature[ranked_features$status == "ok"][[1]]

spec <- model_spec(
  response = assay_feature("taxa", feature),
  subject = subject_var("subject"),
  time = time_var("ageMonths")
)

mf <- model_frame(
  ds,
  trajectory(spec),
  control = trajectory_control(
    engine = "spline",
    transform = "log1p",
    spline_k = 6
  )
)
```

# Inspect grouped folds

```{r folds, eval = FALSE}
folds <- grouped_cv_folds(
  mf$data,
  group = "subject",
  v = 3,
  seed = 1
)

str(folds, max.level = 1)
```

# Run grouped cross-validation

```{r cross-validation, eval = FALSE}
cv <- cross_validate_trajectory(
  mf,
  v = 3,
  group = "subject",
  repeats = 2,
  keep_predictions = TRUE,
  seed = 1
)

cv
```

# Summarize validation performance

```{r validation-output, eval = FALSE}
cv_summary(cv)
head(cv_predictions(cv))
```

# Continue

Use `vignette("feature-testing", package = "LEVAiM")` to apply the same model
operation across multiple assay features.
