---
title: "05. Challenge the trajectory with held-out children"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{05. Challenge the trajectory with held-out children}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(LEVAiM)
```

**Course position:** Lecture 5 of 9<br>
**Previously:** `vignette("spline-fitting", package = "LEVAiM")`<br>
**Next:** `vignette("trajectory-analysis", package = "LEVAiM")`

# Why ordinary row-wise CV is wrong here

If one child contributes samples at birth, four months, and twelve months, a
row-wise split can train on two visits and claim to predict the third. That is
not prediction for a new child. It leaks subject identity and stable biology
across the split.

Our validation unit is therefore the child. Every fold places all samples from
one subject entirely in training or entirely in testing.

# Rebuild the Lecture 4 model frame

```{r model-frame}
input_dir <- file.path("..", "examples", "input_files")
if (!dir.exists(input_dir)) input_dir <- file.path("examples", "input_files")
if (!dir.exists(input_dir)) {
  input_dir <- system.file("extdata", package = "LEVAiM")
}

metadata <- read.csv(
  file.path(input_dir, "2026-07-08-BackhedF_Metadata.csv"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
taxa <- read_metaphlan(
  file.path(
    input_dir,
    "BackhedF_2015_mpa_vOct22_CHOCOPhlAnSGB_202403_lon_subsp_RELAB_merged_profile.tsv"
  ),
  input_type = "merged",
  tax_level = "species",
  include_unclassified = FALSE,
  clean_sample_names = TRUE
)
shared <- intersect(metadata$sample, rownames(taxa))
metadata <- metadata[match(shared, metadata$sample), , drop = FALSE]
rownames(metadata) <- metadata$sample
taxa <- taxa[rownames(metadata), , drop = FALSE]
metadata$subject <- factor(metadata$subject)

ds <- LongitudinalDataset(list(taxa = taxa), metadata, "ageMonths", "subject")
ds <- derive_time_varying_features(
  ds,
  "feeding_state",
  list(feeding_4mo = exposure_window(3, 5, method = "modal"))
)
b_longum <- grep("s__Bifidobacterium_longum$", colnames(taxa), value = TRUE)

spec <- model_spec(
  assay_feature("taxa", b_longum),
  subject_var("subject"),
  time_var("ageMonths"),
  list(group_effect("feeding_4mo", role = "derived_exposure"))
)
mf <- model_frame(
  ds,
  trajectory(spec, varying_trajectory("feeding_4mo")),
  trajectory_control(
    engine = "spline", transform = "log1p", spline_k = 3,
    na_action = "explicit", sparse_min_n = 5
  )
)
```

# Inspect the split before trusting it

```{r grouped-folds}
folds <- grouped_cv_folds(mf$data, group = "subject", v = 5, seed = 7)

fold_audit <- do.call(rbind, lapply(folds, function(fold) {
  train_subjects <- unique(as.character(mf$data$subject[fold$train]))
  test_subjects <- unique(as.character(mf$data$subject[fold$test]))
  data.frame(
    fold = fold$fold,
    train_subjects = length(train_subjects),
    test_subjects = length(test_subjects),
    overlap = length(intersect(train_subjects, test_subjects))
  )
}))
fold_audit
```

Every overlap must be zero. This is a structural test, not a hopeful assumption.

# Run repeated subject-grouped validation

```{r cross-validation}
cv <- cross_validate_trajectory(
  mf,
  v = 5,
  repeats = 2,
  seed = 7,
  keep_predictions = TRUE
)
cv_summary(cv)
```

RMSE and MAE are prediction-scale summaries for the transformed response. They
do not validate the feeding exposure, establish causality, or prove every local
wiggle. They tell us how this specification predicts observations from children
not used to fit that fold.

Here RMSE ranges from about 1.04 to 1.29 and MAE from 0.80 to 0.99 across ten
held-out folds. The spread is reasonably stable, but this lecture has not fit a
null or simpler benchmark model. Stability alone therefore does not establish
good predictive performance.

# Prediction and the null answer different questions

Grouped CV asks whether the fitted specification travels to children who were
not used for estimation. The Lecture 3 critical-period null asks whether the
integrated separation of smooth feeding-group trajectories is unusual under a
reduced model with no focal feeding group, a shared trajectory, and the same
subject structure. The centered-shape null is a different hypothesis and
retains constant group offsets. Neither null substitutes for CV.

The same blocking unit appears in both procedures for different reasons. CV
holds out whole children. The Gaussian trajectory null applies one
wild-bootstrap multiplier to all residuals from a child. Both prevent repeated
observations from masquerading as independent evidence.

```{r null-engine-contract}
trajectory_null_control(
  method = "cluster_wild_bootstrap",
  simulations = 999,
  seed = 7
)
```

This constructs the control object but does not rerun the null fits from
Lecture 3. It makes the validation contract explicit: prediction
error, empirical hypothesis calibration, effect magnitude, and multiplicity
are four separate outputs.

# Inspect errors instead of keeping one score

```{r held-out-predictions}
held_out <- cv_predictions(cv)
head(held_out)

held_out$residual <- held_out$observed - held_out$predicted
aggregate(
  abs(held_out$residual),
  list(repetition = held_out[["repeat"]], fold = held_out$fold),
  mean
)
```

The prediction table is the durable object. It lets us inspect whether errors
concentrate at a particular age, feeding group, fold, or subject instead of
letting one average score close the discussion.

The first rows already show the practical issue: some observed zeros receive
predictions near 2 on the modeled scale. That is a reason to inspect occurrence
separately in Lecture 7, not a detail to average away.

# Pass the baton

Lecture 6 leaves the single-organism view and asks which fitted microbial
trajectories resemble one another.

```{r next-lecture, eval = FALSE}
vignette("trajectory-analysis", package = "LEVAiM")
```
